---
title: "Small Simulations"
format: html
editor: visual
---

```{r}
suppressPackageStartupMessages({
  library(igraph)
  library(Matrix)
  library(dplyr)
  library(ggplot2)
  library(here)
})

source(here::here("R/stats.R"))
source(here::here("R/sampling.R"))
source(here::here("R/simulation.R"))
source(here::here("R/plotting.R"))
```

```{r}
largest_subgraph_undirected <- readRDS(here::here("networks", "largest_subgraph_undirected.rds"))
pathway_stats               <- readRDS(here::here("pathway_info", "pathway_stats.rds"))
bin_cache                   <- readRDS(here::here("networks", "degree_bin_cache.rds"))
```

# Core statistics (ICS norm + Moran mean + Gi\* mean)

Functions: `ics_norm()`, `fast_moran_mean()`, `fast_gistar_mean()`, `compute_stats_fast()` — sourced from `R/stats.R`

# Degree-matched sampling via bin_cache (for ICS null only)

This assumes your bin_cache\[\["bins_4"\]\]\$node_table has columns node and bin, and bin_members is a named list of bin→nodes.

Function: `sample_degree_matched()` — sourced from `R/sampling.R`

# Wrapper A: run one ICS bin setting across the noise sweep (numbers only)

ICS null: degree-binned with num_bins_ics (or uniform if =1)

Moran/Gi\* null: uniform node-set null (as you requested)

returns a data.frame

Function: `run_pathway_noise_sweep()` — sourced from `R/simulation.R`

# Wrapper B: compare multiple ICS bin settings in one run (numbers only)

Function: `run_pathway_noise_sweep_compare_ics_bins()` — sourced from `R/simulation.R`

```{r}
# Single run (ICS bins fixed to 4)
res <- run_pathway_noise_sweep(
  pathway_name = "Cadherin signaling pathway",
  n_total = 100,
  noise_steps = seq(0, 1, by = 0.2),
  n_iterations = 1000,
  num_bins_ics = 4,
  graph = largest_subgraph_undirected,
  pathway_data = pathway_stats,
  bin_cache = bin_cache
)

res
```

```{r}
# Compare ICS bins 1,2,4,6,10 (Moran/Gi* uniform)
res_long <- run_pathway_noise_sweep_compare_ics_bins(
  pathway_name = "Downstream TCR signaling",
  n_total = 100,
  noise_steps = seq(0, 1, by = 0.2),
  n_iterations = 1000,
  ics_bins_set = c(1, 2, 4, 6, 10),
  graph = largest_subgraph_undirected,
  pathway_data = pathway_stats,
  bin_cache = bin_cache
)

res_long

```

```{r}
plot_pvals_clean(res_long)
plot_effect_by_bins(res_long)
```

```{r}
res_long_reg <- run_pathway_noise_sweep_compare_ics_bins(
  pathway_name = "Regulation of insulin secretion",
  n_total = 100,
  noise_steps = seq(0, 1, by = 0.2),
  n_iterations = 1000,
  ics_bins_set = c(1, 2, 4, 6, 10),
  graph = largest_subgraph_undirected,
  pathway_data = pathway_stats,
  bin_cache = bin_cache
)
```

```{r}
plot_pvals_clean(res_long_reg)
plot_effect_by_bins(res_long_reg)
```

```{r}
fp <- false_positive_compare_ics_bins(
  graph = largest_subgraph_undirected,
  bin_cache = bin_cache,
  n_total = 100,
  ics_bins_set = c(1, 2, 4, 6, 10),
  n_outer = 50,        # start smaller for a quick check
  n_iterations = 500,   # then scale up later
  alpha = 0.05,
  seed = 1,
  n_cores      = 7
)

fp

```
```{r}
fp_full <- false_positive_compare_ics_bins(
  graph        = largest_subgraph_undirected,
  bin_cache    = bin_cache,
  n_total      = 100,
  ics_bins_set = c(1, 2, 4, 6, 10),
  n_outer      = 5000,
  n_iterations = 600,
  alpha        = 0.05,
  seed         = 1,
  n_cores      = 7,
  return_pvals = TRUE    # <-- this is what was missing
)

fp_full
```
```{r}
fp_full$pvals %>%
  mutate(fp_bins1 = bins_1 < 0.05,
         fp_bins4 = bins_4 < 0.05) %>%
  group_by(fp_bins1) %>%
  summarise(mean_deg = mean(mean_degree),
            median_deg = mean(median_degree))
```

