---
title: "AUC Single Pathways (Multi-Pathway Comparison)"
format: html
editor: visual
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(igraph)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(foreach)
  library(doParallel)
  library(here)
})

source(here::here("R/stats.R"))
source(here::here("R/sampling.R"))
source(here::here("R/auc.R"))
source(here::here("R/plotting.R"))
```

```{r}
#| label: load-data
largest_subgraph_undirected <- readRDS(here::here("networks", "largest_subgraph_undirected.rds"))
pathway_stats               <- readRDS(here::here("pathway_info", "pathway_stats.rds"))
pw_char                     <- readRDS(here::here("pathway_info", "pathway_characterization.rds"))
bin_cache                   <- readRDS(here::here("networks", "degree_bin_cache.rds"))
nodes                       <- V(largest_subgraph_undirected)$name
```

# Pathway candidate exploration

Select pathways spanning different size and connectivity strata: at least 2 small (20--50 genes), 2 medium (80--200 genes), and 2 large (300+ genes), all with a single connected component.

```{r}
#| label: candidates-small
small_candidates <- pw_char %>%
  filter(n_genes >= 20, n_genes <= 50, n_components == 1) %>%
  arrange(desc(mean_subgraph_deg)) %>%
  head(10)

small_candidates %>%
  select(pathway, n_genes, mean_subgraph_deg, mean_full_deg) %>%
  knitr::kable(digits = 1, caption = "Small pathway candidates (20-50 genes, 1 component)")
```

```{r}
#| label: candidates-medium
medium_candidates <- pw_char %>%
  filter(n_genes >= 80, n_genes <= 200, n_components == 1) %>%
  arrange(desc(mean_subgraph_deg)) %>%
  head(10)

medium_candidates %>%
  select(pathway, n_genes, mean_subgraph_deg, mean_full_deg) %>%
  knitr::kable(digits = 1, caption = "Medium pathway candidates (80-200 genes, 1 component)")
```

```{r}
#| label: candidates-large
large_candidates <- pw_char %>%
  filter(n_genes >= 300, n_components == 1) %>%
  arrange(desc(mean_subgraph_deg)) %>%
  head(10)

large_candidates %>%
  select(pathway, n_genes, mean_subgraph_deg, mean_full_deg) %>%
  knitr::kable(digits = 1, caption = "Large pathway candidates (300+ genes, 1 component)")
```

# Pathway selection

Eight pathways chosen to span size strata and degree profiles.

```{r}
#| label: pathway-selection
pathway_panel <- c(
  "HDR through Homologous Recombination (HRR)",
  "COPI-mediated anterograde transport",
  "Clathrin-mediated endocytosis",
  "Smooth Muscle Contraction",
  "Olfactory Signaling Pathway",
  "Downstream TCR signaling",
  "Antigen activates B Cell Receptor (BCR) leading to generation of second messengers",
  "FCERI mediated Ca+2 mobilization"
)

pathway_labels <- c(
  "HDR through Homologous Recombination (HRR)" = "HRR",
  "COPI-mediated anterograde transport"         = "COPI transport",
  "Clathrin-mediated endocytosis"               = "Clathrin endocytosis",
  "Smooth Muscle Contraction"                   = "Smooth Muscle",
  "Olfactory Signaling Pathway"                 = "Olfactory",
  "Downstream TCR signaling"                    = "TCR signaling",
  "Antigen activates B Cell Receptor (BCR) leading to generation of second messengers"      = "BCR signaling",
  "FCERI mediated Ca+2 mobilization"            = "FCERI Ca2+"
)

# Summary table with characterization stats
pw_summary <- bind_rows(lapply(pathway_panel, function(pw) {
  row <- pathway_stats %>% filter(PATHWAY_NAMES == pw)
  pw_genes <- intersect(unique(unlist(row$Genes)), nodes)
  char_row <- pw_char %>% filter(pathway == pw)
  tibble(
    short_name      = pathway_labels[pw],
    n_genes         = length(pw_genes),
    mean_subgraph_deg = char_row$mean_subgraph_deg,
    mean_full_deg     = char_row$mean_full_deg,
    size_stratum    = case_when(
      length(pw_genes) <= 50  ~ "Small",
      length(pw_genes) <= 200 ~ "Medium",
      TRUE                    ~ "Large"
    )
  )
}))

pw_summary %>%
  knitr::kable(digits = 1, caption = "Selected pathways: size and connectivity profile")
```

# Shared parameters

```{r}
#| label: params
n_total_vec         <- c(100, 200, 500)
signal_noise_levels <- c(0, 20, 40, 60, 80)
reps_per_level      <- 100
n_perm              <- 1000
ics_bins_set        <- 1:10
n_cores             <- 7
```

# KS degree diagnostic

Before any heavy computation, test whether each pathway's genes have a different degree distribution from the full network.

```{r}
#| label: ks-diagnostic
net_degs <- degree(largest_subgraph_undirected)

ks_table <- bind_rows(lapply(pathway_panel, function(pw) {
  row <- pathway_stats %>% filter(PATHWAY_NAMES == pw)
  pw_genes <- intersect(unique(unlist(row$Genes)), nodes)
  pw_degs <- degree(largest_subgraph_undirected, v = pw_genes)
  ks <- ks.test(pw_degs, net_degs)
  tibble(
    pathway         = pathway_labels[pw],
    n_genes         = length(pw_genes),
    mean_degree     = round(mean(pw_degs), 1),
    network_mean_degree = round(mean(net_degs), 1),
    ks_statistic    = round(ks$statistic, 4),
    ks_pvalue       = format.pval(ks$p.value, digits = 3)
  )
}))

ks_table %>%
  knitr::kable(caption = "KS test: pathway gene degree vs network degree")
```

```{r}
#| label: ks-density-plot
#| fig-width: 14
#| fig-height: 8
density_df <- bind_rows(lapply(pathway_panel, function(pw) {
  row <- pathway_stats %>% filter(PATHWAY_NAMES == pw)
  pw_genes <- intersect(unique(unlist(row$Genes)), nodes)
  pw_degs <- degree(largest_subgraph_undirected, v = pw_genes)
  bind_rows(
    tibble(pathway = pathway_labels[pw], source = "Pathway genes", degree = pw_degs),
    tibble(pathway = pathway_labels[pw], source = "Full network", degree = net_degs)
  )
}))

ggplot(density_df, aes(x = degree, fill = source, color = source)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  facet_wrap(~ pathway, nrow = 2, scales = "free_y") +
  scale_x_log10() +
  labs(
    title = "Degree distribution: pathway genes vs full network",
    x     = "Degree (log scale)",
    y     = "Density",
    fill  = NULL, color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

# Feasibility check

At the lowest noise level (`r min(signal_noise_levels)`%), we need `n_total * (1 - min_noise/100)` pathway genes. Check every pathway x n_total combination before running anything.

```{r}
#| label: feasibility
max_signal_frac <- 1 - min(signal_noise_levels) / 100

feasibility <- expand.grid(
  pathway = pathway_panel,
  n_total = n_total_vec,
  stringsAsFactors = FALSE
) %>%
  rowwise() %>%
  mutate(
    pw_short          = pathway_labels[pathway],
    n_genes           = length(intersect(
      unique(unlist(pathway_stats$Genes[pathway_stats$PATHWAY_NAMES == pathway])),
      nodes
    )),
    max_signal_needed = ceiling(n_total * max_signal_frac),
    feasible          = n_genes >= max_signal_needed
  ) %>%
  ungroup()

feasibility %>%
  select(pw_short, n_genes, n_total, max_signal_needed, feasible) %>%
  arrange(pw_short, n_total) %>%
  knitr::kable(caption = paste0(
    "Feasibility: pathway must have >= n_total * ",
    max_signal_frac, " genes (noise levels: ",
    paste(signal_noise_levels, collapse = ", "), "%)"
  ))

cat("\nFeasible combinations:", sum(feasibility$feasible),
    "/", nrow(feasibility), "\n")
if (any(!feasibility$feasible)) {
  cat("SKIPPING:\n")
  feasibility %>%
    filter(!feasible) %>%
    mutate(msg = paste0("  ", pw_short, " @ n_total=", n_total,
                        " (", n_genes, " genes < ", max_signal_needed, " needed)")) %>%
    pull(msg) %>%
    cat(sep = "\n")
}
```

# Run AUC across feasible pathway x n_total combinations

```{r}
#| label: run-all-v2
#| cache: true
feasible_combos <- feasibility %>% filter(feasible)

all_res <- bind_rows(lapply(seq_len(nrow(feasible_combos)), function(i) {
  pw <- feasible_combos$pathway[i]
  nt <- feasible_combos$n_total[i]
  cat(pathway_labels[pw], ", n_total =", nt, "\n")
  res <- run_single_pathway_auc_parallel(
    pathway_name        = pw,
    graph               = largest_subgraph_undirected,
    pathway_data        = pathway_stats,
    bin_cache           = bin_cache,
    n_total             = nt,
    reps_per_level      = reps_per_level,
    signal_noise_levels = signal_noise_levels,
    null_noise_level    = 100,
    n_perm              = n_perm,
    ics_bins_set        = ics_bins_set,
    n_cores             = n_cores
  )
  res %>%
    mutate(score = -log10(pmax(p_value, 1e-300)),
           n_total = nt)
}))
```

# Compute AUC by pathway x n_total x method

```{r}
#| label: compute-auc
auc_tbl <- all_res %>%
  group_by(pathway, n_total, method) %>%
  group_modify(~ {
    ra <- roc_auc_from_p(.x, score_col = "score", label_col = "label")
    tibble(auc = ra$auc, roc = list(ra$roc))
  }) %>%
  ungroup() %>%
  mutate(pw_short = pathway_labels[pathway])

tpr_tbl <- auc_tbl %>%
  transmute(
    pathway, pw_short, n_total, method, auc,
    tpr_at_fpr = map_dbl(roc, ~ max(.x$TPR[.x$FPR <= 0.05], na.rm = TRUE))
  )
```

## Pathway coverage by n_total

```{r}
#| label: coverage
coverage <- all_res %>%
  distinct(pathway, n_total) %>%
  mutate(pw_short = pathway_labels[pathway])

coverage %>%
  mutate(present = TRUE) %>%
  pivot_wider(id_cols = pw_short, names_from = n_total,
              values_from = present, values_fill = FALSE,
              names_prefix = "n_total=") %>%
  knitr::kable(caption = "Pathway presence at each n_total (FALSE = skipped, infeasible)")
```

## AUC summary table

```{r}
#| label: auc-table
auc_tbl %>%
  select(pw_short, n_total, method, auc) %>%
  arrange(pw_short, n_total, desc(auc)) %>%
  knitr::kable(digits = 3,
               caption = "AUC by pathway, sample size, and method")
```

## (a) AUC vs n_total, faceted by pathway

```{r}
#| label: plot-auc-by-pathway
#| fig-width: 16
#| fig-height: 10
ggplot(auc_tbl, aes(x = factor(n_total), y = auc,
                     color = method, group = method)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ pw_short, nrow = 2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "AUC vs sample size, faceted by pathway",
    x     = "n_total",
    y     = "AUC",
    color = "Method"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        strip.text = element_text(face = "bold")) +
  guides(color = guide_legend(ncol = 5))
```

## (b) AUC vs n_total, faceted by method

Which methods are most robust across pathways?

```{r}
#| label: extract-stat-helper
extract_stat <- function(method) {
  case_when(
    grepl("^ICS",         method) ~ "ICS",
    grepl("^GlobalMoran", method) ~ "Global Moran",
    grepl("^Joincount",   method) ~ "Join Count",
    TRUE ~ method
  )
}

extract_bins <- function(method) {
  ifelse(grepl("bins=", method),
         as.integer(sub(".*bins=(\\d+).*", "\\1", method)),
         1L)
}
```

```{r}
#| label: plot-auc-by-method
#| fig-width: 16
#| fig-height: 10
auc_method <- auc_tbl %>%
  mutate(stat = extract_stat(method),
         bins = extract_bins(method)) %>%
  filter(bins %in% c(1, 4, 10))

ggplot(auc_method, aes(x = factor(n_total), y = auc,
                        color = pw_short, group = pw_short)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ method, ncol = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "AUC vs sample size, faceted by method (bins 1, 4, 10)",
    x     = "n_total",
    y     = "AUC",
    color = "Pathway"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8),
        strip.text = element_text(face = "bold", size = 9)) +
  guides(color = guide_legend(ncol = 3))
```

## (c) AUC heatmap at n_total = 200

```{r}
#| label: heatmap-auc
#| fig-width: 14
#| fig-height: 7
heat_df <- auc_tbl %>%
  filter(n_total == 200) %>%
  select(pw_short, method, auc)

ggplot(heat_df, aes(x = method, y = pw_short, fill = auc)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(auc, 2)), size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "#B2D8D8", high = "#004D40",
                       midpoint = 0.7, limits = c(0.4, 1)) +
  labs(
    title = "AUC heatmap: pathway x method (n_total = 200)",
    x     = "Method",
    y     = "Pathway",
    fill  = "AUC"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 10))
```

## Bin-stability: AUC vs bins at n_total = 200

```{r}
#| label: bin-stability
#| fig-width: 16
#| fig-height: 10
bin_stab <- auc_tbl %>%
  filter(n_total == 200) %>%
  mutate(bins = extract_bins(method),
         stat = extract_stat(method))

ggplot(bin_stab, aes(x = bins, y = auc, color = stat, group = stat)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ pw_short, nrow = 2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Bin-stability: AUC vs degree bins (n_total = 200)",
    x     = "Number of degree bins",
    y     = "AUC",
    color = "Method"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

# ROC curves at n_total = 200 (bins 1, 4, 10)

```{r}
#| label: roc-200
#| fig-width: 16
#| fig-height: 10
roc_200 <- auc_tbl %>%
  filter(n_total == 200,
         extract_bins(method) %in% c(1, 4, 10)) %>%
  select(pw_short, method, auc, roc) %>%
  unnest(roc) %>%
  mutate(label = paste0(method, " (", round(auc, 3), ")"))

ggplot(roc_200, aes(x = FPR, y = TPR, color = label)) +
  geom_line(linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  facet_wrap(~ pw_short, nrow = 2) +
  coord_equal() +
  labs(
    title = "ROC curves at n_total = 200 (bins 1, 4, 10)",
    x     = "False Positive Rate",
    y     = "True Positive Rate",
    color = "Method (AUC)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        strip.text = element_text(face = "bold")) +
  guides(color = guide_legend(ncol = 3))
```

# Save results

```{r}
#| label: save
dir.create(here::here("results"), showWarnings = FALSE)
saveRDS(all_res, here::here("results", "single_pathway_auc_raw.rds"))
saveRDS(auc_tbl, here::here("results", "single_pathway_auc_table.rds"))
```

# Summary

```{r}
#| label: summary
# Best method per pathway at n_total = 200
auc_tbl %>%
  filter(n_total == 200) %>%
  group_by(pw_short) %>%
  slice_max(auc, n = 3) %>%
  select(pw_short, method, auc) %>%
  knitr::kable(digits = 3,
               caption = "Top 3 methods per pathway (n_total = 200)")

# Method robustness: mean AUC across pathways at n_total = 200
auc_tbl %>%
  filter(n_total == 200) %>%
  group_by(method) %>%
  summarise(
    mean_auc = mean(auc),
    sd_auc   = sd(auc),
    min_auc  = min(auc),
    .groups  = "drop"
  ) %>%
  arrange(desc(mean_auc)) %>%
  knitr::kable(digits = 3,
               caption = "Method robustness: mean AUC across 8 pathways (n_total = 200)")
```
