---
title: "AUC Multiple Pathways (Mixture Signal)"
format: html
editor: visual
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(igraph)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(foreach)
  library(doParallel)
  library(here)
})

source(here::here("R/stats.R"))
source(here::here("R/sampling.R"))
source(here::here("R/auc.R"))
source(here::here("R/simulation_synthetic.R"))
source(here::here("R/plotting.R"))
```

```{r}
#| label: load-data
largest_subgraph_undirected <- readRDS(here::here("networks", "largest_subgraph_undirected.rds"))
pathway_stats               <- readRDS(here::here("pathway_info", "pathway_stats.rds"))
pw_char                     <- readRDS(here::here("pathway_info", "pathway_characterization.rds"))
bin_cache                   <- readRDS(here::here("networks", "degree_bin_cache.rds"))
```

# Shared parameters

```{r}
#| label: params
K_vec           <- c(100, 200, 500)
signal_frac_vec <- c(0.05, 0.1, 0.2, 0.4, 0.6)
reps            <- 100
n_perm          <- 1000
ics_bins_set    <- 1:10
n_cores         <- 7
```

# Pathway set candidate exploration

Group pathways by degree profile to build sets spanning high, mid, and low degree.

```{r}
#| label: candidates-high
high_deg <- pw_char %>%
  filter(mean_full_deg > 100, n_genes >= 30, n_components == 1) %>%
  arrange(desc(n_genes))

high_deg %>%
  head(15) %>%
  select(pathway, n_genes, mean_subgraph_deg, mean_full_deg) %>%
  knitr::kable(digits = 1, caption = "High-degree candidates (mean_full_deg > 100)")
```

```{r}
#| label: candidates-mid
mid_deg <- pw_char %>%
  filter(mean_full_deg >= 50, mean_full_deg <= 100, n_genes >= 30, n_components == 1) %>%
  arrange(desc(n_genes))

mid_deg %>%
  head(15) %>%
  select(pathway, n_genes, mean_subgraph_deg, mean_full_deg) %>%
  knitr::kable(digits = 1, caption = "Mid-degree candidates (50-100)")
```

```{r}
#| label: candidates-low
low_deg <- pw_char %>%
  filter(mean_full_deg < 50, n_genes >= 30, n_components == 1) %>%
  arrange(desc(n_genes))

low_deg %>%
  head(15) %>%
  select(pathway, n_genes, mean_subgraph_deg, mean_full_deg) %>%
  knitr::kable(digits = 1, caption = "Low-degree candidates (mean_full_deg < 50)")
```

# Pathway set definitions

Seven sets spanning high, mid, low degree profiles, plus a synthetic degree-typical pool.

```{r}
#| label: define-sets
# --- High-degree sets (mean_full_deg > 100) ---
set1_pathways <- c(
  "Downstream TCR signaling",
  "Regulation of insulin secretion",
  "Cadherin signaling pathway"
)
set1_label <- "Set 1: TCR / Insulin / Cadherin"

set2_pathways <- c(
  "HDR through Homologous Recombination (HRR)",
  "Antigen activates B Cell Receptor (BCR) leading to generation of second messengers",
  "Clathrin-mediated endocytosis"
)
set2_label <- "Set 2: HRR / BCR / Clathrin"

set3_pathways <- c(
  "VEGFR2 mediated cell proliferation",
  "Ca2+ pathway",
  "FCERI mediated NF-kB activation"
)
set3_label <- "Set 3: VEGFR2 / Ca2+ / FCERI"

set4_pathways <- c(
  "Signaling by SCF-KIT",
  "DAP12 signaling",
  "Role of phospholipids in phagocytosis"
)
set4_label <- "Set 4: SCF-KIT / DAP12 / Phagocytosis"

# --- Mid-degree set (mean_full_deg 50-100) ---
set5_pathways <- c(
  "Separation of Sister Chromatids",
  "Rho GTPase cycle",
  "Inflammation mediated by chemokine and cytokine signaling pathway"
)
set5_label <- "Set 5: Chromatids / Rho / Inflammation"

# --- Low-degree set (mean_full_deg < 50) ---
set6_pathways <- c(
  "Olfactory Signaling Pathway",
  "Neddylation",
  "Energy dependent regulation of mTOR by LKB1-AMPK"
)
set6_label <- "Set 6: Olfactory / Neddylation / mTOR-AMPK"
```

## Set 7: Synthetic degree-typical pool

500 random genes with degree distribution indistinguishable from the network (KS p \> 0.05). This serves as a degree-typical comparison against the pathway-based pools above.

```{r}
#| label: define-set7
set.seed(777)
set7_result <- select_degree_typical_dispersed(
  largest_subgraph_undirected,
  n_genes = 500, ks_alpha = 0.05, max_attempts = 500, seed = 777
)
set7_genes <- set7_result$genes
cat("Set 7 synthetic pool:", length(set7_genes), "genes\n")
cat("KS p =", round(set7_result$ks_pvalue, 4),
    " | attempts:", set7_result$n_attempts, "\n")

# Create a synthetic pathway entry for compatibility with run_fixed_mixture_auc_parallel
set7_pathway_entry <- data.frame(
  PATHWAY_NAMES = "SYNTHETIC_DEGREE_TYPICAL_500",
  stringsAsFactors = FALSE
)
set7_pathway_entry$Genes <- list(set7_genes)

# Augment pathway_stats with this synthetic entry
pathway_stats_aug <- bind_rows(pathway_stats, set7_pathway_entry)

set7_pathways <- "SYNTHETIC_DEGREE_TYPICAL_500"
set7_label <- "Set 7: Synthetic degree-typical (500 genes)"
```

```{r}
#| label: build-set-list
net_nodes <- V(largest_subgraph_undirected)$name
net_degs  <- degree(largest_subgraph_undirected)

get_pool_genes <- function(pw_names) {
  genes <- lapply(pw_names, function(pw) {
    row <- pathway_stats_aug %>% filter(PATHWAY_NAMES == pw)
    intersect(unique(unlist(row$Genes)), net_nodes)
  })
  unique(unlist(genes))
}

set_list <- list(
  list(label = set1_label, pathways = set1_pathways),
  list(label = set2_label, pathways = set2_pathways),
  list(label = set3_label, pathways = set3_pathways),
  list(label = set4_label, pathways = set4_pathways),
  list(label = set5_label, pathways = set5_pathways),
  list(label = set6_label, pathways = set6_pathways),
  list(label = set7_label, pathways = set7_pathways)
)

# Verify all pathways exist and print pool sizes
for (s in set_list) {
  pool <- get_pool_genes(s$pathways)
  cat(s$label, ": pool =", length(pool), "genes\n")
  missing <- setdiff(s$pathways, pathway_stats_aug$PATHWAY_NAMES)
  if (length(missing) > 0) cat("  WARNING missing:", paste(missing, collapse = ", "), "\n")
}
```

# Degree diagnostic: pool genes vs network

```{r}
#| label: degree-diagnostic
ks_diag <- bind_rows(lapply(set_list, function(s) {
  pool <- get_pool_genes(s$pathways)
  pool_degs <- degree(largest_subgraph_undirected, v = pool)
  ks <- ks.test(pool_degs, net_degs)
  tibble(
    pathway_set      = s$label,
    pool_size        = length(pool),
    pool_mean_deg    = mean(pool_degs),
    network_mean_deg = mean(net_degs),
    pool_median_deg  = median(pool_degs),
    network_median_deg = median(net_degs),
    ks_statistic     = ks$statistic,
    ks_pvalue        = ks$p.value
  )
}))

ks_diag %>%
  knitr::kable(digits = 4,
               caption = "KS test: pool gene degree vs full network degree (all 7 sets)")
```

```{r}
#| label: degree-density-plot
#| fig-width: 16
#| fig-height: 8
density_df <- bind_rows(lapply(set_list, function(s) {
  pool <- get_pool_genes(s$pathways)
  pool_degs <- degree(largest_subgraph_undirected, v = pool)
  bind_rows(
    tibble(pathway_set = s$label, source = "Pool genes", degree = pool_degs),
    tibble(pathway_set = s$label, source = "Full network", degree = net_degs)
  )
}))

ggplot(density_df, aes(x = degree, fill = source, color = source)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  facet_wrap(~ pathway_set, nrow = 2, scales = "free_y") +
  scale_x_log10() +
  labs(
    title = "Degree distribution: pathway pool genes vs full network",
    x     = "Degree (log scale)",
    y     = "Density",
    fill  = NULL, color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

# Feasibility check

A combination is feasible when `pool_size >= K * max(signal_frac_vec)`.

```{r}
#| label: feasibility
feasibility <- bind_rows(lapply(set_list, function(s) {
  pool_size <- length(get_pool_genes(s$pathways))
  expand.grid(K = K_vec, stringsAsFactors = FALSE) %>%
    mutate(
      pathway_set       = s$label,
      pool_size         = pool_size,
      max_signal_needed = ceiling(K * max(signal_frac_vec)),
      feasible          = pool_size >= max_signal_needed
    )
}))

feasibility %>%
  select(pathway_set, pool_size, K, max_signal_needed, feasible) %>%
  arrange(pathway_set, K) %>%
  knitr::kable(caption = paste0(
    "Feasibility: pool_size >= K * max(signal_frac) = K * ",
    max(signal_frac_vec)
  ))

cat("\nFeasible combinations:", sum(feasibility$feasible),
    "/", nrow(feasibility), "\n")
if (any(!feasibility$feasible)) {
  cat("SKIPPING:\n")
  feasibility %>%
    filter(!feasible) %>%
    mutate(msg = paste0("  ", pathway_set, " @ K=", K,
                        " (pool=", pool_size, " < ", max_signal_needed, " needed)")) %>%
    pull(msg) %>%
    cat(sep = "\n")
}
```

# Run all sets

```{r}
#| label: run-all-v2
#| cache: true
all_res <- bind_rows(lapply(set_list, function(s) {
  feasible_K <- feasibility %>%
    filter(pathway_set == s$label, feasible) %>%
    pull(K)

  if (length(feasible_K) == 0) {
    cat("No feasible K for", s$label, "-- skipping.\n")
    return(tibble())
  }

  cat("Running", s$label, "for K =", paste(feasible_K, collapse = ", "), "\n")
  bind_rows(lapply(feasible_K, function(K) {
    cat("  ", s$label, ", K =", K, "\n")
    # Use augmented pathway data for Set 7's synthetic entry
    pw_data <- if (any(s$pathways == "SYNTHETIC_DEGREE_TYPICAL_500")) pathway_stats_aug else pathway_stats
    res <- run_fixed_mixture_auc_parallel(
      graph           = largest_subgraph_undirected,
      pathway_data    = pw_data,
      bin_cache       = bin_cache,
      fixed_pathways  = s$pathways,
      K               = K,
      signal_frac_vec = signal_frac_vec,
      reps            = reps,
      n_perm          = n_perm,
      ics_bins_set    = ics_bins_set,
      n_cores         = n_cores
    )
    res %>% mutate(K = K, pw_set = s$label)
  }))
}))

if (nrow(all_res) == 0) {
  stop("No feasible combinations ran.")
}

cat("Total results:", nrow(all_res), "rows\n")
all_res %>% distinct(pw_set, K) %>% print()
```

# Compute AUC

```{r}
#| label: compute-auc
auc_all <- all_res %>%
  group_by(pw_set, K, signal_frac, method) %>%
  group_modify(~ {
    ra <- roc_auc_from_p(.x, score_col = "score", label_col = "label")
    tibble(auc = ra$auc, roc = list(ra$roc))
  }) %>%
  ungroup()

tpr_all <- auc_all %>%
  transmute(
    pw_set, K, signal_frac, method, auc,
    tpr_at_fpr = map_dbl(roc, ~ max(.x$TPR[.x$FPR <= 0.05], na.rm = TRUE))
  )
```

## Helpers

```{r}
#| label: bin-helpers
extract_stat <- function(method) {
  case_when(
    grepl("^ICS",         method) ~ "ICS",
    grepl("^GlobalMoran", method) ~ "Global Moran",
    grepl("^Joincount",   method) ~ "Join Count",
    TRUE ~ method
  )
}

extract_bins <- function(method) {
  ifelse(grepl("bins=", method),
         as.integer(sub(".*bins=(\\d+).*", "\\1", method)),
         1L)
}
```

# AUC vs signal fraction (all sets, bins 1 and 10)

```{r}
#| label: plot-auc-all-sets
#| fig-width: 16
#| fig-height: 14
auc_ref <- auc_all %>%
  filter(extract_bins(method) %in% c(1, 10))

ggplot(auc_ref, aes(x = signal_frac, y = auc, color = method, group = method)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  facet_grid(pw_set ~ paste0("K = ", K)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(breaks = signal_frac_vec) +
  labs(
    title = "AUC vs signal fraction (bins=1 and bins=10, all sets)",
    x     = "Signal fraction",
    y     = "AUC",
    color = "Method"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        strip.text = element_text(face = "bold", size = 8)) +
  guides(color = guide_legend(ncol = 3))
```

## TPR at FPR=0.05 (K = 200)

```{r}
#| label: plot-tpr-k200
#| fig-width: 16
#| fig-height: 10
tpr_k200 <- tpr_all %>%
  filter(K == 200, extract_bins(method) %in% c(1, 10))

ggplot(tpr_k200, aes(x = signal_frac, y = tpr_at_fpr,
                      color = method, group = method)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ pw_set, nrow = 2) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(breaks = signal_frac_vec) +
  labs(
    title = "TPR at FPR <= 0.05 (K = 200, bins=1 and bins=10)",
    x     = "Signal fraction",
    y     = "TPR @ FPR = 0.05",
    color = "Method"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        strip.text = element_text(face = "bold")) +
  guides(color = guide_legend(ncol = 3))
```

## Bin-stability: AUC vs bins (K = 200, signal_frac = 0.2)

```{r}
#| label: bin-stability
#| fig-width: 16
#| fig-height: 10
bin_stab <- auc_all %>%
  filter(K == 200, signal_frac == 0.2) %>%
  mutate(bins = extract_bins(method),
         stat = extract_stat(method))

ggplot(bin_stab, aes(x = bins, y = auc, color = stat, group = stat)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ pw_set, nrow = 2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Bin-stability: AUC vs degree bins (K = 200, signal = 20%)",
    x     = "Number of degree bins",
    y     = "AUC",
    color = "Method"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

## AUC heatmap at K = 200, signal = 20%

```{r}
#| label: heatmap
#| fig-width: 14
#| fig-height: 8
heat_df <- auc_all %>%
  filter(K == 200, signal_frac == 0.2,
         extract_bins(method) %in% c(1, 4, 10)) %>%
  select(pw_set, method, auc)

ggplot(heat_df, aes(x = method, y = pw_set, fill = auc)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(auc, 2)), size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "#B2D8D8", high = "#004D40",
                       midpoint = 0.7, limits = c(0.4, 1)) +
  labs(
    title = "AUC heatmap: pathway set x method (K = 200, signal = 20%)",
    x     = "Method",
    y     = "Pathway set",
    fill  = "AUC"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

## Global Moran's I: cost of degree control (uniform vs bins=10)

```{r}
#| label: gmoran-cost
gmoran_cost <- auc_all %>%
  filter(method %in% c("GlobalMoran_uniform", "GlobalMoran_dm_bins=10")) %>%
  select(pw_set, K, signal_frac, method, auc) %>%
  pivot_wider(names_from = method, values_from = auc,
              names_prefix = "auc_") %>%
  rename(
    uniform = `auc_GlobalMoran_uniform`,
    bins10  = `auc_GlobalMoran_dm_bins=10`
  ) %>%
  mutate(delta = uniform - bins10)

gmoran_cost %>%
  knitr::kable(digits = 3,
               caption = "Global Moran's I AUC: uniform vs bins=10 (delta = cost of degree control)")
```

## Join Count: cost of degree control (uniform vs bins=10)

```{r}
#| label: jc-cost
jc_cost <- auc_all %>%
  filter(method %in% c("Joincount_uniform", "Joincount_dm_bins=10")) %>%
  select(pw_set, K, signal_frac, method, auc) %>%
  pivot_wider(names_from = method, values_from = auc,
              names_prefix = "auc_") %>%
  rename(
    uniform = `auc_Joincount_uniform`,
    bins10  = `auc_Joincount_dm_bins=10`
  ) %>%
  mutate(delta = uniform - bins10)

jc_cost %>%
  knitr::kable(digits = 3,
               caption = "Join Count AUC: uniform vs bins=10 (delta = cost of degree control)")
```

# Save results

```{r}
#| label: save
dir.create(here::here("results"), showWarnings = FALSE)
saveRDS(all_res, here::here("results", "multi_pathway_auc_raw.rds"))
saveRDS(auc_all, here::here("results", "multi_pathway_auc_table.rds"))
```

# Summary

```{r}
#| label: ks-summary
# KS summary across all 6 sets
ks_diag %>%
  select(pathway_set, pool_size, pool_mean_deg, network_mean_deg, ks_pvalue) %>%
  mutate(hub_enriched = ifelse(ks_pvalue < 0.05, "Yes", "No")) %>%
  knitr::kable(digits = 3,
               caption = "KS summary: which sets are hub-enriched vs degree-typical?")
```

```{r}
#| label: summary-top
# Best methods at K = 200, signal = 20% for each pathway set
auc_all %>%
  filter(K == 200, signal_frac == 0.2) %>%
  group_by(pw_set) %>%
  slice_max(auc, n = 3) %>%
  select(pw_set, method, auc) %>%
  knitr::kable(digits = 3,
               caption = "Top 3 methods per pathway set (K=200, signal=20%)")
```

```{r}
#| label: summary-robustness
# Robustness: mean AUC across all 6 sets at K=200, signal=20%
auc_all %>%
  filter(K == 200, signal_frac == 0.2) %>%
  group_by(method) %>%
  summarise(
    mean_auc = mean(auc),
    sd_auc   = sd(auc),
    min_auc  = min(auc),
    .groups  = "drop"
  ) %>%
  arrange(desc(mean_auc)) %>%
  knitr::kable(digits = 3,
               caption = "Method robustness: mean AUC across all 7 pathway sets (K=200, signal=20%)")
```
