---
title: "Synthetic Simulations (Tier 1)"
format: html
editor: visual
---

```{r}
suppressPackageStartupMessages({
  library(igraph)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(here)
})

source(here::here("R/stats.R"))
source(here::here("R/sampling.R"))
source(here::here("R/simulation_synthetic.R"))
```

```{r}
largest_subgraph_undirected <- readRDS(here::here("networks", "largest_subgraph_undirected.rds"))
bin_cache                   <- readRDS(here::here("networks", "degree_bin_cache.rds"))
```

# Scenario 1: Single BFS cluster

Plant a tight topological cluster of 50 genes via BFS.

```{r}
set.seed(42)
cluster_s1 <- plant_bfs_cluster(largest_subgraph_undirected, cluster_size = 50)
length(cluster_s1)
```

```{r}
n_total_vec <- c(100, 200, 500)

res_s1 <- bind_rows(lapply(n_total_vec, function(nt) {
  run_synthetic_scenario(
    graph        = largest_subgraph_undirected,
    signal_genes = cluster_s1,
    n_total      = nt,
    n_perm       = 500,
    bin_cache    = bin_cache,
    ics_bins_set = c(1, 4, 10),
    seed         = 42
  ) %>% mutate(n_total = nt)
}))

res_s1
```

# Scenario 2: Multiple dispersed BFS clusters

Plant 3 clusters of 20 genes each, separated by \>= 5 hops.

```{r}
set.seed(123)
dispersed <- plant_dispersed_clusters(
  largest_subgraph_undirected,
  n_clusters = 3, cluster_size = 20, min_separation = 5
)

cat("Signal genes:", length(dispersed$signal_genes), "\n")
cat("Seeds:", dispersed$seeds, "\n")
sapply(dispersed$clusters, length)
```

```{r}
res_s2 <- bind_rows(lapply(n_total_vec, function(nt) {
  run_synthetic_scenario(
    graph        = largest_subgraph_undirected,
    signal_genes = dispersed$signal_genes,
    n_total      = nt,
    n_perm       = 500,
    bin_cache    = bin_cache,
    ics_bins_set = c(1, 4, 10),
    seed         = 123
  ) %>% mutate(n_total = nt)
}))

res_s2
```

# Scenario 3: Dispersed hub genes (false positive control)

Select 50 high-degree genes that are NOT topologically clustered. Expect: significant under uniform null, NOT significant under degree-matched null.

```{r}
set.seed(456)
hubs <- select_dispersed_hubs(largest_subgraph_undirected, n_hubs = 50)
cat("Hub genes selected:", length(hubs), "\n")
cat("Mean degree:", mean(degree(largest_subgraph_undirected, v = hubs)), "\n")
```

```{r}
res_s3 <- bind_rows(lapply(n_total_vec, function(nt) {
  run_synthetic_scenario(
    graph        = largest_subgraph_undirected,
    signal_genes = hubs,
    n_total      = nt,
    n_perm       = 500,
    bin_cache    = bin_cache,
    ics_bins_set = c(1, 4, 10),
    seed         = 456
  ) %>% mutate(n_total = nt)
}))

res_s3
```

# Scenario 7: Degree-matched cluster (confound validator)

Plant a BFS cluster whose degree distribution matches the network background.

```{r}
#| include: false
set.seed(789)
deg_matched <- plant_degree_matched_cluster(
  largest_subgraph_undirected,
  cluster_size = 50,
  max_attempts = 100
)

cat("KS test p-value:", deg_matched$ks_pvalue, "\n")
cat("Signal genes:", length(deg_matched$signal_genes), "\n")
```

```{r}
res_s7 <- bind_rows(lapply(n_total_vec, function(nt) {
  run_synthetic_scenario(
    graph        = largest_subgraph_undirected,
    signal_genes = deg_matched$signal_genes,
    n_total      = nt,
    n_perm       = 500,
    bin_cache    = bin_cache,
    ics_bins_set = c(1, 4, 10),
    seed         = 789
  ) %>% mutate(n_total = nt)
}))

res_s7
```

# Summary comparison

```{r}
all_res <- bind_rows(
  res_s1 %>% mutate(scenario = "S1: Single BFS cluster"),
  res_s2 %>% mutate(scenario = "S2: Dispersed clusters"),
  res_s3 %>% mutate(scenario = "S3: Dispersed hubs"),
  res_s7 %>% mutate(scenario = "S7: Degree-matched cluster")
)

all_res %>%
  select(scenario, n_total, method, p_value) %>%
  pivot_wider(names_from = method, values_from = p_value)
```

```{r}
ggplot(all_res, aes(x = method, y = -log10(pmax(p_value, 1e-4)),
                    fill = factor(n_total))) +
  geom_col(position = "dodge") +
  facet_wrap(~ scenario, scales = "free_x") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  labs(
    title = "Synthetic Scenarios: -log10(p) by method and sample size",
    x = "Method", y = "-log10(p-value)", fill = "n_total"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
