---
title: "Leave-One-Out Analysis"
format: html
editor: visual
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(igraph)
  library(dplyr)
  library(ggplot2)
  library(here)
})

source(here::here("R/stats.R"))
source(here::here("R/loo.R"))
source(here::here("R/plotting.R"))
```

```{r}
#| label: load-data
largest_subgraph_undirected <- readRDS(here::here("networks", "largest_subgraph_undirected.rds"))
pathway_stats               <- readRDS(here::here("pathway_info", "pathway_stats.rds"))
```

# Validate loo_gene() against brute-force

Run both the optimized (articulation-point) and naive (brute-force) versions on a small pathway and confirm deltas match to floating-point tolerance.

```{r}
#| label: validation-setup
# Pick a small pathway for validation (~20-30 genes)
small_pw <- pathway_stats %>%
  mutate(n_on_net = sapply(Genes, function(g) length(intersect(unique(unlist(g)), V(largest_subgraph_undirected)$name)))) %>%
  filter(n_on_net >= 15, n_on_net <= 40) %>%
  slice(1)

val_nodes <- intersect(unique(unlist(small_pw$Genes)), V(largest_subgraph_undirected)$name)
cat("Validation pathway:", small_pw$PATHWAY_NAMES, "\n")
cat("Genes on network:", length(val_nodes), "\n")
```

```{r}
#| label: validation-run
loo_fast  <- loo_gene(largest_subgraph_undirected, val_nodes)
loo_naive <- loo_gene_naive(largest_subgraph_undirected, val_nodes)

# Merge and compare
merged <- merge(
  loo_fast[, c("gene", "delta")],
  loo_naive[, c("gene", "delta")],
  by = "gene", suffixes = c("_fast", "_naive")
)

max_diff <- max(abs(merged$delta_fast - merged$delta_naive), na.rm = TRUE)
cat("Max absolute difference:", max_diff, "\n")
stopifnot(max_diff < 1e-12)
cat("VALIDATION PASSED: loo_gene() matches brute-force within floating-point tolerance.\n")
```

# Single pathway LOO: Downstream TCR signaling

```{r}
#| label: single-pathway-setup
pathway_name <- "Downstream TCR signaling"

row <- pathway_stats %>% filter(PATHWAY_NAMES == pathway_name)
tcr_nodes <- intersect(unique(unlist(row$Genes)), V(largest_subgraph_undirected)$name)
cat("Pathway:", pathway_name, "\n")
cat("Genes on network:", length(tcr_nodes), "\n")
```

## Component-level LOO

```{r}
#| label: single-comp-loo
comp_loo <- loo_component(largest_subgraph_undirected, tcr_nodes)
comp_loo %>% select(comp_id, comp_size, ics_full, ics_without, delta) %>% arrange(desc(abs(delta)))
```

```{r}
#| label: single-comp-plot
plot_loo_components(comp_loo, title = paste0("Component LOO: ", pathway_name))
```

```{r}
#| label: single-comp-summary
cat("Total components:", nrow(comp_loo), "\n")
cat("Largest component:", max(comp_loo$comp_size), "genes\n")
cat("Components with |delta| > 0.01:", sum(abs(comp_loo$delta) > 0.01), "\n")
```

## Gene-level LOO

```{r}
#| label: single-gene-loo
gene_loo <- loo_gene(largest_subgraph_undirected, tcr_nodes)
```

```{r}
#| label: single-gene-summary
cat("Total genes:", nrow(gene_loo), "\n")
cat("Articulation points:", sum(gene_loo$is_articulation), "\n")
cat("Non-articulation (analytical):", sum(!gene_loo$is_articulation), "\n")
```

```{r}
#| label: single-gene-plot
plot_loo_importance_vs_contribution(gene_loo, title = paste0("Gene LOO: ", pathway_name))
```

```{r}
#| label: single-gene-cor
cor_val <- cor(gene_loo$subgraph_degree, gene_loo$delta,
               use = "complete.obs", method = "spearman")
cat("Spearman correlation (degree vs delta):", round(cor_val, 3), "\n")
```

## Degree diagnostic

```{r}
#| label: single-degree
plot_degree_diagnostic(largest_subgraph_undirected, tcr_nodes,
                       title = paste0("Degree distribution: ", pathway_name, " vs background"))
```

```{r}
#| label: single-degree-stats
deg_all <- degree(largest_subgraph_undirected)
deg_pw  <- degree(largest_subgraph_undirected, v = tcr_nodes)

cat("Network mean degree:", round(mean(deg_all), 2), "\n")
cat("Pathway mean degree:", round(mean(deg_pw), 2), "\n")
cat("KS test p-value:", ks.test(deg_pw, deg_all)$p.value, "\n")
```

## Top genes tables

```{r}
#| label: single-top-delta
# Top 10 by |delta| (highest contribution to ICS)
gene_loo %>%
  arrange(desc(abs(delta))) %>%
  select(gene, subgraph_degree, is_articulation, comp_size, delta) %>%
  head(10)
```

```{r}
#| label: single-top-degree
# Top 10 by subgraph degree (most connected within pathway)
gene_loo %>%
  arrange(desc(subgraph_degree)) %>%
  select(gene, subgraph_degree, is_articulation, comp_size, delta) %>%
  head(10)
```

```{r}
#| label: single-bridges
# High contribution but low importance (interesting bridges)
gene_loo %>%
  filter(subgraph_degree <= median(subgraph_degree, na.rm = TRUE)) %>%
  arrange(desc(abs(delta))) %>%
  select(gene, subgraph_degree, is_articulation, comp_size, delta) %>%
  head(10)
```

# Multi-pathway LOO: 3 pathways + noise

Combine genes from 3 pathways and add random noise genes, then run LOO to examine which genes and components drive the ICS signal in a realistic multi-source gene set.

```{r}
#| label: multi-setup
multi_pathways <- c(
  "Downstream TCR signaling",
  "Regulation of insulin secretion",
  "Cadherin signaling pathway"
)

nodes <- V(largest_subgraph_undirected)$name

# Collect genes per pathway
pw_gene_lists <- lapply(multi_pathways, function(pw) {
  row <- pathway_stats %>% filter(PATHWAY_NAMES == pw)
  intersect(unique(unlist(row$Genes)), nodes)
})
names(pw_gene_lists) <- multi_pathways

for (pw in multi_pathways) {
  cat(pw, ":", length(pw_gene_lists[[pw]]), "genes on network\n")
}

# Union of pathway genes
signal_pool <- unique(unlist(pw_gene_lists))
cat("\nUnion of pathway genes:", length(signal_pool), "\n")

# Add noise: 30% of total set size
set.seed(42)
n_noise <- as.integer(round(length(signal_pool) * 0.3))
noise_genes <- sample(setdiff(nodes, signal_pool), n_noise)
cat("Noise genes:", n_noise, "\n")

multi_target <- c(signal_pool, noise_genes)
cat("Total multi-pathway set:", length(multi_target), "\n")
```

```{r}
#| label: multi-origin
# Label each gene by its origin (pathway 1, 2, 3, multi, or noise)
gene_origin <- data.frame(gene = multi_target, stringsAsFactors = FALSE)
gene_origin$origin <- sapply(gene_origin$gene, function(g) {
  in_pw <- sapply(pw_gene_lists, function(gl) g %in% gl)
  n_pw <- sum(in_pw)
  if (n_pw == 0) return("Noise")
  if (n_pw > 1) return("Multi-pathway")
  names(which(in_pw))
})
table(gene_origin$origin)
```

## Multi-pathway component-level LOO

```{r}
#| label: multi-comp-loo
comp_loo_multi <- loo_component(largest_subgraph_undirected, multi_target)
comp_loo_multi %>%
  select(comp_id, comp_size, ics_full, ics_without, delta) %>%
  arrange(desc(abs(delta)))
```

```{r}
#| label: multi-comp-plot
plot_loo_components(comp_loo_multi, title = "Component LOO: 3-pathway + noise mixture")
```

```{r}
#| label: multi-comp-composition
# For each component, what fraction of genes come from each origin?
for (i in seq_len(nrow(comp_loo_multi))) {
  cg <- comp_loo_multi$comp_genes[[i]]
  origins <- gene_origin$origin[match(cg, gene_origin$gene)]
  if (length(cg) >= 3) {
    cat("Component", comp_loo_multi$comp_id[i],
        "(n=", comp_loo_multi$comp_size[i],
        ", delta=", round(comp_loo_multi$delta[i], 4), "):",
        paste(names(table(origins)), table(origins), sep = "=", collapse = ", "), "\n")
  }
}
```

## Multi-pathway gene-level LOO

```{r}
#| label: multi-gene-loo
gene_loo_multi <- loo_gene(largest_subgraph_undirected, multi_target)
```

```{r}
#| label: multi-gene-summary
cat("Total genes:", nrow(gene_loo_multi), "\n")
cat("Articulation points:", sum(gene_loo_multi$is_articulation), "\n")
cat("Non-articulation:", sum(!gene_loo_multi$is_articulation), "\n")
```

```{r}
#| label: multi-gene-plot
#| fig-width: 10
#| fig-height: 7
# Annotate with origin
gene_loo_multi <- gene_loo_multi %>%
  left_join(gene_origin, by = "gene")

ggplot(gene_loo_multi, aes(x = subgraph_degree, y = delta, color = origin)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE,
              color = "grey40", linetype = "dashed") +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  labs(
    title = "Gene LOO: 3-pathway + noise mixture",
    subtitle = paste0("Spearman r = ",
                      round(cor(gene_loo_multi$subgraph_degree,
                                gene_loo_multi$delta,
                                use = "complete.obs", method = "spearman"), 3)),
    x     = "Subgraph degree",
    y     = "LOO delta",
    color = "Gene origin"
  ) +
  theme_minimal(base_size = 13)
```

## LOO delta by origin

Do pathway genes contribute more to ICS than noise genes?

```{r}
#| label: multi-delta-by-origin
gene_loo_multi %>%
  group_by(origin) %>%
  summarise(
    n         = n(),
    mean_delta = mean(delta, na.rm = TRUE),
    sd_delta   = sd(delta, na.rm = TRUE),
    n_art      = sum(is_articulation),
    mean_degree = mean(subgraph_degree),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_delta)) %>%
  knitr::kable(digits = 4,
               caption = "LOO delta summary by gene origin")
```

```{r}
#| label: multi-delta-boxplot
#| fig-width: 10
#| fig-height: 5
ggplot(gene_loo_multi, aes(x = origin, y = delta, fill = origin)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  labs(
    title = "LOO delta distribution by gene origin",
    x     = "Origin",
    y     = "LOO delta (contribution to ICS)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Top contributing genes (multi-pathway)

```{r}
#| label: multi-top-delta
gene_loo_multi %>%
  arrange(desc(abs(delta))) %>%
  select(gene, origin, subgraph_degree, is_articulation, comp_size, delta) %>%
  head(15) %>%
  knitr::kable(digits = 4,
               caption = "Top 15 genes by |delta| in 3-pathway mixture")
```

```{r}
#| label: multi-noise-contribution
# What fraction of top-20 |delta| genes are noise?
top20 <- gene_loo_multi %>%
  arrange(desc(abs(delta))) %>%
  head(20)
cat("Noise genes in top 20 by |delta|:", sum(top20$origin == "Noise"), "/", 20, "\n")
cat("Multi-pathway genes in top 20:", sum(top20$origin == "Multi-pathway"), "/", 20, "\n")
```

# Summary

```{r}
#| label: summary
cat("=== Single pathway LOO (", pathway_name, ") ===\n")
cat("Components:", nrow(comp_loo), "\n")
cat("Articulation points:", sum(gene_loo$is_articulation), "/", nrow(gene_loo), "\n")
cat("Degree-delta Spearman r:", round(cor_val, 3), "\n\n")

cat("=== Multi-pathway LOO (3 pathways + 30% noise) ===\n")
cat("Total genes:", nrow(gene_loo_multi), "\n")
cat("Components:", nrow(comp_loo_multi), "\n")
cat("Articulation points:", sum(gene_loo_multi$is_articulation), "/", nrow(gene_loo_multi), "\n")
cat("Degree-delta Spearman r:",
    round(cor(gene_loo_multi$subgraph_degree, gene_loo_multi$delta,
              use = "complete.obs", method = "spearman"), 3), "\n")

# Signal vs noise separation
signal_delta <- mean(gene_loo_multi$delta[gene_loo_multi$origin != "Noise"], na.rm = TRUE)
noise_delta  <- mean(gene_loo_multi$delta[gene_loo_multi$origin == "Noise"], na.rm = TRUE)
cat("Mean delta (signal genes):", round(signal_delta, 5), "\n")
cat("Mean delta (noise genes):", round(noise_delta, 5), "\n")
```
