---
title: "Synthetic Scenario AUC Analysis"
format: html
editor: visual
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(igraph)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(foreach)
  library(doParallel)
  library(here)
})

source(here::here("R/stats.R"))
source(here::here("R/sampling.R"))
source(here::here("R/simulation_synthetic.R"))
source(here::here("R/auc.R"))
source(here::here("R/plotting.R"))
```

```{r}
#| label: load-data
largest_subgraph_undirected <- readRDS(here::here("networks", "largest_subgraph_undirected.rds"))
bin_cache                   <- readRDS(here::here("networks", "degree_bin_cache.rds"))
```

# Shared parameters

```{r}
#| label: params
n_total_vec    <- c(100, 200, 500)
reps_per_level <- 100
n_perm         <- 1000
ics_bins_set   <- 1:10
n_cores        <- 7
```

# AUC engine

For each replicate: creates an alternative gene set (planted signal + noise) and a null gene set (pure random), computes p-values for all methods under shared degree-matched null, then computes AUC across replicates.

```{r}
#| label: auc-engine
run_synthetic_auc_parallel <- function(
  graph,
  signal_genes,
  n_total_vec,
  reps_per_level = 50,
  n_perm         = 500,
  bin_cache,
  ics_bins_set   = c(1, 4, 10),
  n_cores        = 7,
  scenario_name  = "Synthetic",
  seed           = 1
) {
  prep  <- prep_graph_fast(graph)
  g     <- prep$g
  A     <- prep$A
  nodes <- prep$nodes

  signal_genes <- intersect(signal_genes, nodes)
  background   <- setdiff(nodes, signal_genes)

  # Task grid: n_total × rep × label (1 = alternative, 0 = null)
  task_df <- expand.grid(
    n_total = n_total_vec,
    rep     = seq_len(reps_per_level),
    label   = c(1L, 0L),
    stringsAsFactors = FALSE
  ) %>% mutate(task_id = row_number())

  cl <- parallel::makeCluster(n_cores)
  on.exit(parallel::stopCluster(cl), add = TRUE)
  doParallel::registerDoParallel(cl)

  parallel::clusterEvalQ(cl, {
    suppressPackageStartupMessages({ library(igraph); library(Matrix) })
    NULL
  })

  parallel::clusterExport(
    cl,
    varlist = c(
      "g", "A", "nodes", "signal_genes", "background",
      "n_perm", "ics_bins_set", "seed", "bin_cache", "scenario_name",
      "ics_norm", "fast_joincount", "global_moran_I",
      "sample_degree_matched"
    ),
    envir = environment()
  )

  res_long <- foreach(
    ii = seq_len(nrow(task_df)),
    .combine = dplyr::bind_rows,
    .packages = c("igraph", "Matrix", "dplyr")
  ) %dopar% {

    nt     <- task_df$n_total[ii]
    label  <- task_df$label[ii]
    rep_id <- task_df$rep[ii]

    set.seed(seed + ii)

    n_sig   <- min(length(signal_genes), nt)
    n_noise <- nt - n_sig

    if (label == 1L) {
      # Alternative: planted signal + random noise
      target_nodes <- c(
        sample(signal_genes, n_sig),
        if (n_noise > 0) sample(background, n_noise) else character(0)
      )
    } else {
      # Null: pure random
      target_nodes <- sample(nodes, nt, replace = FALSE)
    }

    idx_target <- match(target_nodes, nodes)
    x_obs <- integer(length(nodes)); x_obs[idx_target] <- 1L

    obs_ics <- ics_norm(induced_subgraph(g, vids = target_nodes))
    obs_gm  <- global_moran_I(A, x_obs)
    obs_jc  <- fast_joincount(A, x_obs)

    # Degree-matched null per bin scheme — shared sample for ALL methods
    all_out <- list()

    for (b in ics_bins_set) {
      rand_ics <- numeric(n_perm)
      rand_gm  <- numeric(n_perm)
      rand_jc  <- numeric(n_perm)

      for (p in seq_len(n_perm)) {
        sset <- if (b > 1) {
          sample_degree_matched(target_nodes, bin_cache,
                                num_bins = b, replace = TRUE)
        } else {
          sample(nodes, nt, replace = TRUE)
        }
        idx_s <- match(sset, nodes)
        x_s   <- integer(length(nodes)); x_s[idx_s] <- 1L

        rand_ics[p] <- ics_norm(induced_subgraph(g, vids = sset))
        rand_gm[p]  <- global_moran_I(A, x_s)
        rand_jc[p]  <- fast_joincount(A, x_s)
      }

      null_label <- if (b > 1) paste0("dm_bins=", b) else "uniform"

      all_out[[length(all_out) + 1]] <- c(
        setNames(mean(rand_ics <= obs_ics), paste0("ICS_bins=", b)),
        setNames(mean(rand_gm >= obs_gm),   paste0("GlobalMoran_", null_label)),
        setNames(mean(rand_jc >= obs_jc),   paste0("Joincount_", null_label))
      )
    }

    out <- unlist(all_out)

    data.frame(
      scenario = scenario_name,
      n_total  = nt,
      rep      = rep_id,
      label    = label,
      method   = names(out),
      p_value  = as.numeric(out),
      stringsAsFactors = FALSE
    )
  }

  res_long %>% mutate(score = -log10(pmax(p_value, 1e-300)))
}
```

```{r}
#| label: auc-helpers
compute_auc_table <- function(res_long) {
  res_long %>%
    group_by(scenario, n_total, method) %>%
    group_modify(~ {
      ra <- roc_auc_from_p(.x, score_col = "score", label_col = "label")
      tibble(auc = ra$auc, roc = list(ra$roc))
    }) %>%
    ungroup()
}

compute_tpr_table <- function(auc_tbl, fpr_target = 0.05) {
  auc_tbl %>%
    transmute(
      scenario, n_total, method, auc,
      tpr_at_fpr = map_dbl(roc, ~ max(.x$TPR[.x$FPR <= fpr_target], na.rm = TRUE))
    ) %>%
    arrange(scenario, n_total, desc(tpr_at_fpr))
}
```

# Plant signal sets

```{r}
#| label: plant-signals
set.seed(42)
signal_s1 <- plant_bfs_cluster(largest_subgraph_undirected, cluster_size = 50)
cat("S1 signal genes:", length(signal_s1), "\n")
subg_s1 <- induced_subgraph(largest_subgraph_undirected, vids = signal_s1)
cat("S1 components:", components(subg_s1)$no,
    " | density:", round(edge_density(subg_s1), 4), "\n")

set.seed(123)
dispersed_s2 <- plant_dispersed_clusters(
  largest_subgraph_undirected,
  n_clusters = 3, cluster_size = 20, min_separation = 5
)
signal_s2 <- dispersed_s2$signal_genes
cat("S2 signal genes:", length(signal_s2),
    " | seeds:", dispersed_s2$seeds,
    " | cluster sizes:", sapply(dispersed_s2$clusters, length), "\n")

set.seed(789)
deg_matched_s7 <- plant_degree_matched_cluster(
  largest_subgraph_undirected, cluster_size = 50, max_attempts = 200
)
signal_s7 <- deg_matched_s7$signal_genes
cat("S7 signal genes:", length(signal_s7),
    " | KS p =", round(deg_matched_s7$ks_pvalue, 4), "\n")

set.seed(456)
hubs_s3 <- select_dispersed_hubs(largest_subgraph_undirected, n_hubs = 50)
cat("S3 hub genes:", length(hubs_s3),
    " | mean degree:", round(mean(degree(largest_subgraph_undirected, v = hubs_s3)), 1), "\n")

# S8: Degree-typical dispersed (GWAS-like: small clusters + singletons)
s8_result <- plant_degree_typical_dispersed(
  largest_subgraph_undirected,
  n_signal = 50, n_small_clusters = 3, cluster_size = c(5, 8),
  ks_alpha = 0.05, max_attempts = 500, seed = 888
)
signal_s8 <- s8_result$signal_genes
cat("S8 signal genes:", length(signal_s8),
    " | clusters:", length(s8_result$clusters),
    " | singletons:", length(s8_result$singletons),
    " | KS p =", round(s8_result$ks_pvalue, 4),
    " | attempts:", s8_result$n_attempts, "\n")

# S3b: Degree-typical dispersed null (no clustering)
s3b_result <- select_degree_typical_dispersed(
  largest_subgraph_undirected,
  n_genes = 50, ks_alpha = 0.05, max_attempts = 500, seed = 999
)
signal_s3b <- s3b_result$genes
cat("S3b null genes:", length(signal_s3b),
    " | KS p =", round(s3b_result$ks_pvalue, 4),
    " | attempts:", s3b_result$n_attempts, "\n")
```

## KS degree diagnostics for all signal sets

```{r}
#| label: ks-diagnostics
net_degs <- degree(largest_subgraph_undirected)

s1_degs <- degree(largest_subgraph_undirected, v = signal_s1)
s1_ks <- ks.test(s1_degs, net_degs)
cat("S1 (single BFS cluster): mean degree =", round(mean(s1_degs), 1),
    ", KS p =", format.pval(s1_ks$p.value, digits = 3), "\n")

s2_degs <- degree(largest_subgraph_undirected, v = signal_s2)
s2_ks <- ks.test(s2_degs, net_degs)
cat("S2 (dispersed clusters): mean degree =", round(mean(s2_degs), 1),
    ", KS p =", format.pval(s2_ks$p.value, digits = 3), "\n")

s7_degs <- degree(largest_subgraph_undirected, v = signal_s7)
s7_ks <- ks.test(s7_degs, net_degs)
cat("S7 (degree-matched cluster): mean degree =", round(mean(s7_degs), 1),
    ", KS p =", format.pval(s7_ks$p.value, digits = 3), "\n")

cat("S3 (dispersed hubs): mean degree =",
    round(mean(degree(largest_subgraph_undirected, v = hubs_s3)), 1), "\n")

s8_degs <- degree(largest_subgraph_undirected, v = signal_s8)
s8_ks <- ks.test(s8_degs, net_degs)
cat("S8 (degree-typical dispersed): mean degree =", round(mean(s8_degs), 1),
    ", KS p =", format.pval(s8_ks$p.value, digits = 3), "\n")

s3b_degs <- degree(largest_subgraph_undirected, v = signal_s3b)
s3b_ks <- ks.test(s3b_degs, net_degs)
cat("S3b (degree-typical null): mean degree =", round(mean(s3b_degs), 1),
    ", KS p =", format.pval(s3b_ks$p.value, digits = 3), "\n")

ks_diag <- tibble(
  scenario     = c("S1: Single BFS", "S2: Dispersed", "S7: Degree-matched",
                   "S3: Hubs", "S8: Degree-typical dispersed", "S3b: Degree-typical null"),
  n_genes      = c(length(signal_s1), length(signal_s2), length(signal_s7),
                   length(hubs_s3), length(signal_s8), length(signal_s3b)),
  mean_degree  = c(mean(s1_degs), mean(s2_degs), mean(s7_degs),
                   mean(degree(largest_subgraph_undirected, v = hubs_s3)),
                   mean(s8_degs), mean(s3b_degs)),
  network_mean = mean(net_degs),
  ks_pvalue    = c(s1_ks$p.value, s2_ks$p.value, s7_ks$p.value,
                   ks.test(degree(largest_subgraph_undirected, v = hubs_s3), net_degs)$p.value,
                   s8_ks$p.value, s3b_ks$p.value)
)

ks_diag %>%
  mutate(mean_degree = round(mean_degree, 1),
         network_mean = round(network_mean, 1),
         ks_pvalue = format.pval(ks_pvalue, digits = 3)) %>%
  knitr::kable(caption = "KS test: signal gene degree vs network degree")
```

# S1: Single BFS cluster — AUC

50 signal genes planted as one tight connected cluster via BFS.

```{r}
#| label: s1-auc-v2
#| cache: true
res_s1 <- run_synthetic_auc_parallel(
  graph          = largest_subgraph_undirected,
  signal_genes   = signal_s1,
  n_total_vec    = n_total_vec,
  reps_per_level = reps_per_level,
  n_perm         = n_perm,
  bin_cache      = bin_cache,
  ics_bins_set   = ics_bins_set,
  n_cores        = n_cores,
  scenario_name  = "S1: Single cluster",
  seed           = 100
)
```

# S2: Dispersed clusters — AUC

3 clusters of 20 genes each, BFS-planted with minimum 5-hop separation.

```{r}
#| label: s2-auc-v2
#| cache: true
res_s2 <- run_synthetic_auc_parallel(
  graph          = largest_subgraph_undirected,
  signal_genes   = signal_s2,
  n_total_vec    = n_total_vec,
  reps_per_level = reps_per_level,
  n_perm         = n_perm,
  bin_cache      = bin_cache,
  ics_bins_set   = ics_bins_set,
  n_cores        = n_cores,
  scenario_name  = "S2: Dispersed clusters",
  seed           = 200
)
```

# S7: Degree-matched cluster — AUC

50 signal genes in a BFS cluster whose degree distribution approximates the network background.

```{r}
#| label: s7-auc-v2
#| cache: true
res_s7 <- run_synthetic_auc_parallel(
  graph          = largest_subgraph_undirected,
  signal_genes   = signal_s7,
  n_total_vec    = n_total_vec,
  reps_per_level = reps_per_level,
  n_perm         = n_perm,
  bin_cache      = bin_cache,
  ics_bins_set   = ics_bins_set,
  n_cores        = n_cores,
  scenario_name  = "S7: Degree-matched cluster",
  seed           = 700
)
```

# S8: Degree-typical dispersed — AUC

GWAS-like signal: a few small BFS clusters plus isolated singletons, all degree-typical (KS p \> 0.05). Tests whether methods can detect mild, realistic clustering when degree is not confounded.

```{r}
#| label: s8-auc
#| cache: true
res_s8 <- run_synthetic_auc_parallel(
  graph          = largest_subgraph_undirected,
  signal_genes   = signal_s8,
  n_total_vec    = n_total_vec,
  reps_per_level = reps_per_level,
  n_perm         = n_perm,
  bin_cache      = bin_cache,
  ics_bins_set   = ics_bins_set,
  n_cores        = n_cores,
  scenario_name  = "S8: Degree-typical dispersed",
  seed           = 800
)
```

# Combined AUC results

```{r}
#| label: combine-results
all_res <- bind_rows(res_s1, res_s2, res_s7, res_s8)
auc_tbl <- compute_auc_table(all_res)
tpr_tbl <- compute_tpr_table(auc_tbl, fpr_target = 0.05)
```

## AUC and TPR summary

```{r}
#| label: auc-summary
auc_tbl %>%
  left_join(tpr_tbl %>% select(scenario, n_total, method, tpr_at_fpr),
            by = c("scenario", "n_total", "method")) %>%
  select(scenario, n_total, method, auc, tpr_at_fpr) %>%
  arrange(scenario, n_total, desc(auc)) %>%
  knitr::kable(digits = 3,
               caption = "AUC and TPR@FPR=0.05 by scenario, sample size, and method")
```

## (a) AUC vs n_total, faceted by scenario

```{r}
#| label: plot-auc-vs-ntotal
#| fig-width: 14
#| fig-height: 6
ggplot(auc_tbl, aes(x = factor(n_total), y = auc,
                     color = method, group = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ scenario, nrow = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Detection power: AUC by method and sample size",
    subtitle = "Degree-matched nulls for all methods; dashed = random classifier",
    x        = "n_total (signal + noise genes)",
    y        = "AUC",
    color    = "Method"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 9),
        strip.text = element_text(face = "bold")) +
  guides(color = guide_legend(ncol = 3))
```

## TPR at FPR = 0.05 vs n_total

```{r}
#| label: plot-tpr-vs-ntotal
#| fig-width: 14
#| fig-height: 6
ggplot(tpr_tbl, aes(x = factor(n_total), y = tpr_at_fpr,
                     color = method, group = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ scenario, nrow = 1) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Detection power: TPR at FPR = 0.05",
    subtitle = "How often each method detects real signal while controlling false positives",
    x        = "n_total",
    y        = "TPR @ FPR = 0.05",
    color    = "Method"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 9),
        strip.text = element_text(face = "bold")) +
  guides(color = guide_legend(ncol = 3))
```

## (b) ROC curves at n_total = 200

```{r}
#| label: plot-roc-200
#| fig-width: 14
#| fig-height: 5
roc_200 <- auc_tbl %>%
  filter(n_total == 200) %>%
  select(scenario, method, auc, roc) %>%
  unnest(roc) %>%
  mutate(label = paste0(method, " (", round(auc, 3), ")"))

ggplot(roc_200, aes(x = FPR, y = TPR, color = label)) +
  geom_line(linewidth = 0.9) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  facet_wrap(~ scenario, nrow = 1) +
  coord_equal() +
  labs(
    title = "ROC curves at n_total = 200",
    x     = "False Positive Rate",
    y     = "True Positive Rate",
    color = "Method (AUC)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8),
        strip.text = element_text(face = "bold")) +
  guides(color = guide_legend(ncol = 3))
```

### ROC zoom: FPR 0–0.1

```{r}
#| label: plot-roc-zoom
#| fig-width: 14
#| fig-height: 5
ggplot(roc_200, aes(x = FPR, y = TPR, color = label)) +
  geom_line(linewidth = 0.9) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0.05, linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~ scenario, nrow = 1) +
  coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 1)) +
  labs(
    title    = "ROC curves zoomed (n_total = 200, FPR 0-0.1)",
    subtitle = "Dotted line = FPR = 0.05",
    x        = "False Positive Rate",
    y        = "True Positive Rate",
    color    = "Method (AUC)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8),
        strip.text = element_text(face = "bold")) +
  guides(color = guide_legend(ncol = 3))
```

# S3: Dispersed hubs — False positive rate

S3 is a false positive control: high-degree dispersed genes with no topological clustering. We run 200 replicates and measure what fraction of p-values fall below 0.05 for each method × bins combination.

```{r}
#| label: s3-fpr-engine
run_hub_fpr_parallel <- function(
  graph,
  hub_genes,
  n_total,
  n_reps       = 200,
  n_perm       = 500,
  bin_cache,
  ics_bins_set = c(1, 4, 10),
  seed         = 1,
  n_cores      = 7
) {
  prep  <- prep_graph_fast(graph)
  g     <- prep$g
  A     <- prep$A
  nodes <- prep$nodes

  hub_genes  <- intersect(hub_genes, nodes)
  background <- setdiff(nodes, hub_genes)
  n_hub      <- min(length(hub_genes), n_total)
  n_noise    <- n_total - n_hub

  cl <- parallel::makeCluster(n_cores)
  on.exit(parallel::stopCluster(cl), add = TRUE)
  doParallel::registerDoParallel(cl)

  parallel::clusterEvalQ(cl, {
    suppressPackageStartupMessages({ library(igraph); library(Matrix) })
    NULL
  })

  parallel::clusterExport(
    cl,
    varlist = c(
      "g", "A", "nodes", "hub_genes", "background",
      "n_total", "n_hub", "n_noise", "n_perm",
      "ics_bins_set", "seed",
      "fast_joincount", "global_moran_I", "ics_norm",
      "sample_degree_matched", "bin_cache"
    ),
    envir = environment()
  )

  foreach(
    rr = seq_len(n_reps),
    .combine = dplyr::bind_rows,
    .packages = c("igraph", "Matrix", "dplyr")
  ) %dopar% {

    set.seed(seed + rr)

    # Hub-enriched gene set — not truly clustered, so rejection = false positive
    target_nodes <- c(
      sample(hub_genes, n_hub),
      if (n_noise > 0) sample(background, n_noise) else character(0)
    )

    idx_target <- match(target_nodes, nodes)
    x_obs <- integer(length(nodes)); x_obs[idx_target] <- 1L

    # KS test: does this hub-enriched set's degree differ from network?
    target_degs  <- degree(g, v = target_nodes)
    network_degs <- degree(g)
    ks_p <- ks.test(target_degs, network_degs)$p.value

    obs_ics <- ics_norm(induced_subgraph(g, vids = target_nodes))
    obs_gm  <- global_moran_I(A, x_obs)
    obs_jc  <- fast_joincount(A, x_obs)

    all_out <- list()

    for (b in ics_bins_set) {
      rand_ics <- numeric(n_perm)
      rand_gm  <- numeric(n_perm)
      rand_jc  <- numeric(n_perm)

      for (p in seq_len(n_perm)) {
        sset <- if (b > 1) {
          sample_degree_matched(target_nodes, bin_cache,
                                num_bins = b, replace = TRUE)
        } else {
          sample(nodes, n_total, replace = TRUE)
        }
        idx_s <- match(sset, nodes)
        x_s   <- integer(length(nodes)); x_s[idx_s] <- 1L

        rand_ics[p] <- ics_norm(induced_subgraph(g, vids = sset))
        rand_gm[p]  <- global_moran_I(A, x_s)
        rand_jc[p]  <- fast_joincount(A, x_s)
      }

      null_label <- if (b > 1) paste0("dm_bins=", b) else "uniform"

      all_out[[length(all_out) + 1]] <- c(
        setNames(mean(rand_ics <= obs_ics), paste0("ICS_bins=", b)),
        setNames(mean(rand_gm >= obs_gm),   paste0("GlobalMoran_", null_label)),
        setNames(mean(rand_jc >= obs_jc),   paste0("Joincount_", null_label))
      )
    }

    out <- unlist(all_out)

    data.frame(
      rep      = rr,
      method   = names(out),
      p_value  = as.numeric(out),
      ks_pvalue = ks_p,
      stringsAsFactors = FALSE
    )
  }
}
```

```{r}
#| label: s3-run-v2
#| cache: true
fpr_s3_raw <- bind_rows(lapply(n_total_vec, function(nt) {
  cat("S3 FPR, n_total =", nt, "\n")
  res <- run_hub_fpr_parallel(
    graph        = largest_subgraph_undirected,
    hub_genes    = hubs_s3,
    n_total      = nt,
    n_reps       = 200,
    n_perm       = n_perm,
    bin_cache    = bin_cache,
    ics_bins_set = ics_bins_set,
    seed         = 300 + nt,
    n_cores      = n_cores
  )
  res %>% mutate(n_total = nt)
}))
```

## S3: False positive rate table

A well-calibrated method under degree-matched null should have FPR near 0.05. Inflated FPR under uniform null (bins=1) is expected because hubs create apparent clustering via degree confound.

```{r}
#| label: s3-fpr-table
fpr_table <- fpr_s3_raw %>%
  group_by(n_total, method) %>%
  summarise(
    n_reps   = n(),
    n_reject = sum(p_value <= 0.05),
    fpr      = mean(p_value <= 0.05),
    mean_p   = mean(p_value),
    .groups  = "drop"
  ) %>%
  arrange(n_total, method)

fpr_table %>%
  select(n_total, method, fpr, mean_p) %>%
  pivot_wider(names_from = n_total, values_from = c(fpr, mean_p),
              names_glue = "{.value}_{n_total}") %>%
  select(method, starts_with("fpr"), starts_with("mean_p")) %>%
  knitr::kable(digits = 3,
               caption = "S3 (dispersed hubs): False positive rate at alpha=0.05 (200 reps)")
```

```{r}
#| label: plot-s3-fpr
#| fig-width: 12
#| fig-height: 6
ggplot(fpr_table, aes(x = method, y = fpr, fill = factor(n_total))) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  labs(
    title    = "S3: False positive rate for hub genes (no real clustering)",
    subtitle = "Red dashed = nominal alpha = 0.05. Bars above = degree confound not controlled.",
    x        = "Method",
    y        = "False positive rate",
    fill     = "n_total"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_legend(ncol = 3))
```

## S3: KS test diagnostic — does the test detect hub enrichment?

The KS test compares each replicate's degree distribution against the full network. If hub enrichment creates a detectable degree shift, KS p \< 0.05 flags it. The hub fraction varies across n_total (50 hubs in 100 vs 50 in 500), so we break down by n_total.

```{r}
#| label: s3-ks-summary
ks_df <- fpr_s3_raw %>%
  distinct(n_total, rep, ks_pvalue)

ks_summary <- ks_df %>%
  group_by(n_total) %>%
  summarise(
    n_reps          = n(),
    hub_frac        = round(min(length(hubs_s3), first(n_total)) / first(n_total), 2),
    frac_ks_reject  = mean(ks_pvalue < 0.05),
    mean_ks_p       = mean(ks_pvalue),
    median_ks_p     = median(ks_pvalue),
    min_ks_p        = min(ks_pvalue),
    max_ks_p        = max(ks_pvalue),
    .groups = "drop"
  )

ks_summary %>%
  knitr::kable(digits = 4,
               caption = "S3: KS test summary per n_total (degree of hub-enriched set vs network)")
```

```{r}
#| label: plot-s3-ks-hist
#| fig-width: 12
#| fig-height: 4
ggplot(ks_df, aes(x = ks_pvalue)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", linewidth = 0.8) +
  facet_wrap(~ paste0("n_total = ", n_total, "\n(hub frac = ",
                       round(pmin(length(hubs_s3), n_total) / n_total, 2), ")"),
             nrow = 1, scales = "free_y") +
  labs(
    title    = "S3: KS p-value distribution (hub-enriched set degree vs network degree)",
    subtitle = "Red dashed = 0.05. Mass at left = KS reliably detects hub enrichment.",
    x        = "KS p-value",
    y        = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))
```

## S3: P-value distribution (calibration diagnostic)

Well-calibrated methods should produce uniform p-value distributions under the null.

```{r}
#| label: plot-s3-pval-dist
#| fig-width: 14
#| fig-height: 10
fpr_s3_raw %>%
  filter(n_total == 200) %>%
  ggplot(aes(x = p_value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  facet_wrap(~ method, scales = "free_y", ncol = 3) +
  labs(
    title    = "S3: P-value distribution under hub-enriched null (n_total = 200)",
    subtitle = "Well-calibrated methods show uniform distribution; pile-up at 0 = false positives",
    x        = "p-value",
    y        = "Count"
  ) +
  theme_minimal(base_size = 11)
```

# S3b: Degree-typical null — False positive rate

S3b is a calibration control: degree-typical genes (KS p \> 0.05) with no topological clustering. A well-calibrated test should reject at approximately the nominal alpha = 0.05. FPR substantially below 0.05 would indicate the degree-matched null is over-conservative for degree-typical gene sets, suppressing true positives. FPR near 0.05 confirms proper calibration.

```{r}
#| label: s3b-fpr-run
#| cache: true
fpr_s3b_raw <- bind_rows(lapply(n_total_vec, function(nt) {
  cat("S3b FPR, n_total =", nt, "\n")
  res <- run_hub_fpr_parallel(
    graph        = largest_subgraph_undirected,
    hub_genes    = signal_s3b,
    n_total      = nt,
    n_reps       = 200,
    n_perm       = n_perm,
    bin_cache    = bin_cache,
    ics_bins_set = ics_bins_set,
    seed         = 3000,
    n_cores      = n_cores
  )
  res %>% mutate(n_total = nt)
}))
```

## S3b: False positive rate table

Degree-typical genes should show well-calibrated FPR under ALL null models (both uniform and degree-matched), since there is no degree confound.

```{r}
#| label: s3b-fpr-table
fpr_table_s3b <- fpr_s3b_raw %>%
  group_by(n_total, method) %>%
  summarise(
    n_reps   = n(),
    n_reject = sum(p_value <= 0.05),
    fpr      = mean(p_value <= 0.05),
    mean_p   = mean(p_value),
    .groups  = "drop"
  ) %>%
  arrange(n_total, method)

fpr_table_s3b %>%
  select(n_total, method, fpr, mean_p) %>%
  pivot_wider(names_from = n_total, values_from = c(fpr, mean_p),
              names_glue = "{.value}_{n_total}") %>%
  select(method, starts_with("fpr"), starts_with("mean_p")) %>%
  knitr::kable(digits = 3,
               caption = "S3b (degree-typical null): False positive rate at alpha=0.05 (200 reps)")
```

## S3 vs S3b: Side-by-side FPR comparison

```{r}
#| label: s3b-vs-s3-comparison
comparison <- bind_rows(
  fpr_table_s3b %>% mutate(scenario = "S3b: Degree-typical null"),
  fpr_table %>% mutate(scenario = "S3: Dispersed hubs")
) %>%
  filter(grepl("bins=10|uniform|bins=1$", method)) %>%
  select(scenario, n_total, method, fpr) %>%
  pivot_wider(names_from = scenario, values_from = fpr)

comparison %>%
  knitr::kable(digits = 3,
               caption = "FPR comparison: hubs vs degree-typical null (key methods)")
```

```{r}
#| label: plot-s3b-fpr
#| fig-width: 12
#| fig-height: 6
fpr_table_s3b %>%
  filter(n_total == 200) %>%
  mutate(
    bins = as.integer(gsub(".*bins=(\\d+).*", "\\1", method)),
    bins = ifelse(grepl("uniform|bins=1$", method), 1L, bins),
    stat = case_when(
      grepl("^GlobalMoran", method) ~ "Global Moran's I",
      grepl("^ICS", method) ~ "ICS",
      grepl("^Joincount", method) ~ "Join count"
    )
  ) %>%
  filter(!is.na(stat)) %>%
  ggplot(aes(x = factor(bins), y = fpr, fill = stat)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(
    title = "S3b calibration: FPR for degree-typical null (n = 200)",
    subtitle = "Dashed red line = nominal alpha = 0.05",
    x    = "Number of degree bins",
    y    = "False positive rate",
    fill = "Method"
  ) +
  theme_minimal(base_size = 13)
```

## S3b: P-value distribution (calibration diagnostic)

```{r}
#| label: plot-s3b-pval-dist
#| fig-width: 14
#| fig-height: 10
fpr_s3b_raw %>%
  filter(n_total == 200) %>%
  ggplot(aes(x = p_value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  facet_wrap(~ method, scales = "free_y", ncol = 3) +
  labs(
    title    = "S3b: P-value distribution under degree-typical null (n_total = 200)",
    subtitle = "Well-calibrated methods show uniform distribution",
    x        = "p-value",
    y        = "Count"
  ) +
  theme_minimal(base_size = 11)
```

# Bin-stability summary

For each detection scenario, plot AUC vs bins (1–10) for all methods at n_total = 200. This parallels the bin-stability curves from notebook 11 (GWAS CRC real data).

```{r}
#| label: bin-stability-extract
# Extract the bins value from method names
extract_bins <- function(method) {
  # ICS_bins=4 -> 4; Moran_dm_bins=4 -> 4; Moran_uniform -> 1; etc.
  ifelse(grepl("bins=", method),
         as.integer(sub(".*bins=(\\d+).*", "\\1", method)),
         1L)
}

extract_stat <- function(method) {
  case_when(
    grepl("^ICS",         method) ~ "ICS",
    grepl("^GlobalMoran", method) ~ "Global Moran",
    grepl("^Joincount",   method) ~ "Join Count",
    TRUE ~ method
  )
}
```

## Detection scenarios: AUC vs bins

```{r}
#| label: bin-stability-auc
#| fig-width: 14
#| fig-height: 6
bin_auc <- auc_tbl %>%
  filter(n_total == 200) %>%
  mutate(
    bins = extract_bins(method),
    stat = extract_stat(method)
  )

ggplot(bin_auc, aes(x = bins, y = auc, color = stat, group = stat)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ scenario, nrow = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Bin-stability: AUC vs number of degree bins (n_total = 200)",
    subtitle = "Parallels bin-stability from notebook 11 (GWAS CRC real data)",
    x        = "Number of degree bins",
    y        = "AUC",
    color    = "Method"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

## Detection scenarios: TPR\@FPR=0.05 vs bins

```{r}
#| label: bin-stability-tpr
#| fig-width: 14
#| fig-height: 6
bin_tpr <- tpr_tbl %>%
  filter(n_total == 200) %>%
  mutate(
    bins = extract_bins(method),
    stat = extract_stat(method)
  )

ggplot(bin_tpr, aes(x = bins, y = tpr_at_fpr, color = stat, group = stat)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ scenario, nrow = 1) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Bin-stability: TPR@FPR=0.05 vs degree bins (n_total = 200)",
    x     = "Number of degree bins",
    y     = "TPR @ FPR = 0.05",
    color = "Method"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

## S3: FPR vs bins

```{r}
#| label: bin-stability-fpr
#| fig-width: 14
#| fig-height: 6
bin_fpr <- fpr_table %>%
  filter(n_total == 200) %>%
  mutate(
    bins = extract_bins(method),
    stat = extract_stat(method)
  )

ggplot(bin_fpr, aes(x = bins, y = fpr, color = stat, group = stat)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "S3 bin-stability: FPR vs degree bins (n_total = 200)",
    subtitle = "Well-calibrated methods stay near 0.05 (red dashed) across all bin counts",
    x        = "Number of degree bins",
    y        = "False positive rate",
    color    = "Method"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

# Save results

```{r}
#| label: save
dir.create(here::here("results"), showWarnings = FALSE)

saveRDS(all_res,      here::here("results", "synthetic_auc_raw.rds"))
saveRDS(auc_tbl,      here::here("results", "synthetic_auc_table.rds"))
saveRDS(tpr_tbl,      here::here("results", "synthetic_tpr_table.rds"))
saveRDS(fpr_s3_raw,   here::here("results", "synthetic_s3_fpr_raw.rds"))
saveRDS(fpr_table,    here::here("results", "synthetic_s3_fpr_summary.rds"))
saveRDS(fpr_s3b_raw,  here::here("results", "synthetic_s3b_fpr_raw.rds"))
saveRDS(fpr_table_s3b, here::here("results", "synthetic_s3b_fpr_summary.rds"))
```

# Summary

```{r}
#| label: summary
cat("=== Detection scenarios: AUC summary ===\n\n")
auc_tbl %>%
  left_join(tpr_tbl %>% select(scenario, n_total, method, tpr_at_fpr),
            by = c("scenario", "n_total", "method")) %>%
  select(scenario, n_total, method, auc, tpr_at_fpr) %>%
  group_by(scenario, method) %>%
  summarise(
    auc_100 = auc[n_total == 100],
    auc_200 = auc[n_total == 200],
    auc_500 = auc[n_total == 500],
    tpr_200 = tpr_at_fpr[n_total == 200],
    .groups = "drop"
  ) %>%
  arrange(scenario, desc(auc_200)) %>%
  knitr::kable(digits = 3,
               caption = "AUC across sample sizes and TPR@FPR=0.05 at n=200")

cat("\n=== S3 false positive control (n_total=200) ===\n\n")
fpr_table %>%
  filter(n_total == 200) %>%
  select(method, fpr, mean_p) %>%
  arrange(method) %>%
  knitr::kable(digits = 3,
               caption = "S3 FPR at n=200 (expect ~0.05 for degree-matched, inflated for uniform)")
```

Key questions answered:

1.  **Detection power (S1, S2, S7):** Which methods best distinguish real clustering from random, and how does power change with sample size?
2.  **Multi-cluster advantage (S2 vs S1):** Does ICS's component-counting logic give it relative advantage when signal is dispersed across multiple modules?
3.  **Degree confound (S7):** When signal genes have typical degree, do degree-controlled methods retain power while uncorrected methods lose it?
4.  **False positive control (S3):** Which methods correctly identify hub-driven apparent clustering as spurious under each null model?
