---
title: "GWAS CRC Real-Data Analysis"
format: html
editor: visual
---

Real-data analysis of CRC GWAS genes on the integrated network. This is what the synthetic simulations are benchmarking against.

```{r}
suppressPackageStartupMessages({
  library(igraph)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(RColorBrewer)
  library(scales)
  library(ggraph)
  library(here)
})

source(here::here("R/stats.R"))
source(here::here("R/sampling.R"))
source(here::here("R/simulation.R"))
source(here::here("R/plotting.R"))
```

```{r}
largest_subgraph_undirected <- readRDS(here::here("networks", "largest_subgraph_undirected.rds"))
bin_cache                   <- readRDS(here::here("networks", "degree_bin_cache.rds"))
gwas_genes_crc              <- readRDS(here::here("data", "gwas_genes_crc.rds"))
```

```{r}
# Filter to genes actually on the network
gwas_on_net <- intersect(gwas_genes_crc, V(largest_subgraph_undirected)$name)
cat("GWAS genes total:", length(gwas_genes_crc), "\n")
cat("GWAS genes on network:", length(gwas_on_net), "\n")
```

# Degree diagnostic

```{r}
plot_degree_diagnostic(largest_subgraph_undirected, gwas_on_net,
                       title = "CRC GWAS genes vs network background")
```

```{r}
deg_all <- degree(largest_subgraph_undirected)
node_names <- V(largest_subgraph_undirected)$name

df_deg <- data.frame(
  gene   = node_names,
  degree = deg_all,
  group  = ifelse(node_names %in% gwas_on_net, "GWAS genes", "Background")
)

ggplot(df_deg, aes(x = degree, fill = group)) +
  geom_density(alpha = 0.4) +
  coord_cartesian(xlim = c(0, quantile(deg_all, 0.95))) +
  labs(
    title = "Degree distribution (linear scale)",
    x     = "Degree",
    y     = "Density",
    fill  = ""
  ) +
  theme_minimal(base_size = 13)
```

```{r}
deg_gwas <- degree(largest_subgraph_undirected, v = gwas_on_net)

cat("Network mean degree:", round(mean(deg_all), 2), "\n")
cat("GWAS mean degree:", round(mean(deg_gwas), 2), "\n")
cat("GWAS median degree:", median(deg_gwas), "\n")
cat("KS test p-value:", format.pval(ks.test(deg_gwas, deg_all)$p.value), "\n")
```

# GWAS subnetwork visualization

```{r}
#| label: gwas-subnetwork-simple
#| fig-width: 10
#| fig-height: 10

gwas_subgraph <- induced_subgraph(largest_subgraph_undirected, vids = gwas_on_net)
comp <- components(gwas_subgraph)

set.seed(42)
ggraph(gwas_subgraph, layout = "fr") +
  geom_edge_link(alpha = 0.4) +
  geom_node_point(color = "darkblue", size = 4) +
  geom_node_text(aes(label = name), repel = TRUE, size = 2.5, max.overlaps = 20) +
  theme_void() +
  ggtitle(paste0("CRC GWAS subnetwork (", length(gwas_on_net), " genes, ",
                  comp$no, " components)"))
```

```{r}
#| label: gwas-subnetwork-annotated
#| fig-width: 10
#| fig-height: 10

# Build node metadata
node_df <- data.frame(
  name = V(gwas_subgraph)$name,
  comp_id = comp$membership,
  comp_size = comp$csize[comp$membership]
)

# Assign colors: largest component = red, other multi-gene = Set2 palette, singletons = grey
largest_comp_id <- which.max(comp$csize)
node_df$color <- "grey70"
node_df$color[node_df$comp_id == largest_comp_id] <- "#E63946"

multi_comp_ids <- setdiff(
  unique(node_df$comp_id[node_df$comp_size > 1]),
  largest_comp_id
)
if (length(multi_comp_ids) > 0) {
  palette <- RColorBrewer::brewer.pal(max(length(multi_comp_ids), 3), "Set2")
  for (i in seq_along(multi_comp_ids)) {
    node_df$color[node_df$comp_id == multi_comp_ids[i]] <- palette[i]
  }
}

V(gwas_subgraph)$color <- node_df$color

set.seed(42)
ggraph(gwas_subgraph, layout = "fr") +
  geom_edge_link(alpha = 0.4) +
  geom_node_point(aes(color = color), size = 4, show.legend = FALSE) +
  scale_color_identity() +
  geom_node_text(aes(label = name), repel = TRUE, size = 2.5, max.overlaps = 20) +
  theme_void() +
  ggtitle(paste0("CRC GWAS subnetwork — colored by component (",
                  comp$no, " components, largest = ", max(comp$csize), " genes)"))
```

# Permutation tests: all methods, bins 1-10

Wrap the GWAS gene list into a pathway_data structure so we can use the existing simulation function.

```{r}
gwas_pathway_data <- data.frame(
  PATHWAY_NAMES = "CRC_GWAS",
  stringsAsFactors = FALSE
)
gwas_pathway_data$Genes <- list(gwas_on_net)
```

```{r}
res_bins <- run_pathway_noise_sweep_compare_ics_bins(
  pathway_name   = "CRC_GWAS",
  n_total        = length(gwas_on_net),
  noise_steps    = 0,
  n_iterations   = 1000,
  ics_bins_set   = 1:10,
  graph          = largest_subgraph_undirected,
  pathway_data   = gwas_pathway_data,
  bin_cache      = bin_cache,
  seed           = 42,
  replace_null   = TRUE,
  run_uniform_null = TRUE
)
```

## Degree-matched null p-values by bin scheme

```{r}
res_dm <- res_bins %>%
  filter(!grepl("sensitivity", null_type))

res_dm %>%
  select(method, null_type, ics_bins, observed, null_mean, null_sd, p_value) %>%
  arrange(method, ics_bins)
```

## Uniform null p-values (sensitivity)

```{r}
res_uniform <- res_bins %>%
  filter(grepl("sensitivity", null_type))

if (nrow(res_uniform) > 0) {
  res_uniform %>%
    select(method, null_type, observed, null_mean, null_sd, p_value)
}
```

## Bin-stability curve: all methods

How p-values change as the number of degree-matching bins varies from 1 to 10.

```{r}
# Extract stat type from method name for grouping
stability_df <- res_dm %>%
  mutate(
    stat = case_when(
      grepl("^ICS",         method) ~ "ICS",
      grepl("^Global Moran", method) ~ "Global Moran's I",
      grepl("^Local Moran",  method) ~ "Local Moran's I",
      grepl("^Getis",       method) ~ "Gi*",
      grepl("^Join",        method) ~ "Join count",
      TRUE ~ method
    )
  )

ggplot(stability_df, aes(x = ics_bins, y = p_value, color = stat, group = stat)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "grey40") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    title = "Bin-stability curve: CRC GWAS genes",
    subtitle = "p-value vs number of degree-matching bins (1 = uniform, 2+ = degree-matched)",
    x     = "Number of degree bins",
    y     = "One-sided p-value",
    color = "Method"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
# Same plot on -log10 scale for clearer separation at small p-values
ggplot(stability_df, aes(x = ics_bins, y = -log10(pmax(p_value, 1e-4)),
                         color = stat, group = stat)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Bin-stability curve: CRC GWAS genes (-log10 scale)",
    x     = "Number of degree bins",
    y     = "-log10(p-value)",
    color = "Method"
  ) +
  theme_minimal(base_size = 13)
```

## Effect size stability

```{r}
ggplot(stability_df, aes(x = ics_bins, y = effect_directional,
                         color = stat, group = stat)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Effect size stability: CRC GWAS genes",
    x     = "Number of degree bins",
    y     = "Directional effect (observed - null mean)",
    color = "Method"
  ) +
  theme_minimal(base_size = 13)
```

# Structural decomposition: GWAS subgraph

Analyse the GWAS-induced subgraph structure directly — connected components, articulation points, and bridge genes — without ICS-based LOO.

```{r}
#| label: gwas-subgraph
gwas_subg <- induced_subgraph(largest_subgraph_undirected, vids = gwas_on_net)
comp <- components(gwas_subg)

cat("GWAS genes on network:", length(gwas_on_net), "\n")
cat("Connected components:", comp$no, "\n")
cat("Largest component size:", max(comp$csize), "\n")
cat("Isolated genes (degree 0 in subgraph):", sum(degree(gwas_subg) == 0), "\n")
cat("Edge density:", round(edge_density(gwas_subg), 5), "\n")
cat("ICS (normalised):", round(ics_norm(gwas_subg), 5), "\n")
```

## Component size distribution

```{r}
#| label: comp-sizes
comp_df <- tibble(
  comp_id   = seq_along(comp$csize),
  comp_size = comp$csize
) %>%
  arrange(desc(comp_size))

comp_df %>%
  count(comp_size, name = "n_components") %>%
  arrange(desc(comp_size)) %>%
  knitr::kable(caption = "GWAS subgraph: component size distribution")
```

```{r}
#| label: plot-comp-sizes
#| fig-width: 10
#| fig-height: 5
comp_df_top <- comp_df %>%
  filter(comp_size > 1) %>%
  mutate(comp_label = paste0("C", comp_id, " (n=", comp_size, ")"))

ggplot(comp_df_top, aes(x = reorder(comp_label, -comp_size), y = comp_size)) +
  geom_col(fill = "steelblue", width = 0.7) +
  labs(
    title = "GWAS subgraph: connected components (size > 1)",
    x     = "Component",
    y     = "Number of genes"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Articulation points

Articulation (cut) points are genes whose removal increases the number of connected components in the GWAS subgraph — they are structural bridges holding clusters together.

```{r}
#| label: articulation-points
art_points <- articulation_points(gwas_subg)
art_names <- V(gwas_subg)$name[art_points]

cat("Articulation points:", length(art_names), "/", length(gwas_on_net),
    "(", round(100 * length(art_names) / length(gwas_on_net), 1), "%)\n")

# For each articulation point: subgraph degree and component membership
art_df <- tibble(
  gene            = V(gwas_subg)$name,
  subgraph_degree = degree(gwas_subg),
  full_degree     = degree(largest_subgraph_undirected, v = V(gwas_subg)$name),
  comp_id         = comp$membership,
  comp_size       = comp$csize[comp$membership],
  is_articulation = gene %in% art_names
)

art_df %>%
  filter(is_articulation) %>%
  arrange(desc(subgraph_degree)) %>%
  select(gene, subgraph_degree, full_degree, comp_size) %>%
  knitr::kable(caption = "Articulation points in GWAS subgraph (bridge genes)")
```

```{r}
#| label: plot-art-vs-degree
#| fig-width: 10
#| fig-height: 6
ggplot(art_df, aes(x = subgraph_degree, y = full_degree, color = is_articulation)) +
  geom_point(alpha = 0.6, size = 2.5) +
  scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick")) +
  labs(
    title = "GWAS genes: subgraph degree vs full network degree",
    x     = "Degree within GWAS subgraph",
    y     = "Degree in full network",
    color = "Articulation point"
  ) +
  theme_minimal(base_size = 13)
```

## What happens when articulation points are removed?

For each articulation point, compute how many new components are created by its removal.

```{r}
#| label: art-point-impact
art_impact <- bind_rows(lapply(art_names, function(g) {
  subg_minus <- induced_subgraph(gwas_subg, vids = setdiff(V(gwas_subg)$name, g))
  new_comps <- components(subg_minus)$no
  tibble(
    gene              = g,
    subgraph_degree   = degree(gwas_subg, v = g),
    comps_without     = new_comps,
    delta_components  = new_comps - comp$no
  )
})) %>%
  arrange(desc(delta_components), desc(subgraph_degree))

art_impact %>%
  knitr::kable(caption = "Articulation point impact: new components created by removal")
```

```{r}
#| label: plot-art-impact
#| fig-width: 10
#| fig-height: 5
ggplot(art_impact, aes(x = reorder(gene, -delta_components), y = delta_components)) +
  geom_col(fill = "firebrick", width = 0.7) +
  labs(
    title = "Articulation point impact: additional components on removal",
    x     = "Gene",
    y     = "Additional components created"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Top connected genes in GWAS subgraph

```{r}
#| label: top-connected
art_df %>%
  arrange(desc(subgraph_degree)) %>%
  select(gene, subgraph_degree, full_degree, is_articulation, comp_size) %>%
  head(20) %>%
  knitr::kable(caption = "Top 20 GWAS genes by subgraph degree")
```

## Component-Level Leave-One-Out

For each connected component in the GWAS-induced subgraph, remove its member genes from the GWAS gene set and rerun Global Moran's I with bins = 10 degree-stratified permutation null (same parameters as the main test). `delta_p = p_value(without component) - p_value(full set)` quantifies each component's contribution to the global signal.

```{r}
#| label: component-loo
#| cache: true

library(foreach)
library(doParallel)

n_perm_loo <- 1000
n_cores    <- 7

prep_loo  <- prep_graph_fast(largest_subgraph_undirected)
g_loo     <- prep_loo$g
A_loo     <- prep_loo$A
nodes_loo <- prep_loo$nodes

# Baseline: full-set Global Moran's I p-value (bins = 10)
idx_full <- match(gwas_on_net, nodes_loo)
x_full   <- integer(length(nodes_loo)); x_full[idx_full] <- 1L
obs_gm_full <- global_moran_I(A_loo, x_full)

set.seed(42)
rand_full <- numeric(n_perm_loo)
for (i in seq_len(n_perm_loo)) {
  s     <- sample_degree_matched(gwas_on_net, bin_cache, num_bins = 10, replace = TRUE)
  idx_s <- match(s, nodes_loo)
  x_s   <- integer(length(nodes_loo)); x_s[idx_s] <- 1L
  rand_full[i] <- global_moran_I(A_loo, x_s)
}
p_value_full <- mean(rand_full >= obs_gm_full)

# Component LOO in parallel
membership_vec <- comp$membership
names(membership_vec) <- V(gwas_subg)$name
unique_comp_ids <- sort(unique(membership_vec))

cl_comp <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl_comp)
parallel::clusterEvalQ(cl_comp, {
  suppressPackageStartupMessages({ library(igraph); library(Matrix) })
  NULL
})
parallel::clusterExport(
  cl_comp,
  varlist = c("A_loo", "nodes_loo", "bin_cache", "n_perm_loo",
              "gwas_on_net", "membership_vec",
              "global_moran_I", "sample_degree_matched"),
  envir = environment()
)

component_loo_raw <- foreach(cid = unique_comp_ids, .combine = bind_rows) %dopar% {
  comp_genes <- names(membership_vec)[membership_vec == cid]
  reduced    <- setdiff(gwas_on_net, comp_genes)

  if (length(reduced) < 3) {
    return(data.frame(
      comp_id         = cid,
      comp_size       = length(comp_genes),
      comp_genes_raw  = I(list(comp_genes)),
      p_value_without = NA_real_,
      stringsAsFactors = FALSE
    ))
  }

  idx_obs <- match(reduced, nodes_loo)
  x_obs   <- integer(length(nodes_loo)); x_obs[idx_obs] <- 1L
  obs_gm  <- global_moran_I(A_loo, x_obs)

  set.seed(42L + as.integer(cid))
  rand_gm <- numeric(n_perm_loo)
  for (i in seq_len(n_perm_loo)) {
    s     <- sample_degree_matched(reduced, bin_cache, num_bins = 10, replace = TRUE)
    idx_s <- match(s, nodes_loo)
    x_s   <- integer(length(nodes_loo)); x_s[idx_s] <- 1L
    rand_gm[i] <- global_moran_I(A_loo, x_s)
  }

  data.frame(
    comp_id         = cid,
    comp_size       = length(comp_genes),
    comp_genes_raw  = I(list(comp_genes)),
    p_value_without = mean(rand_gm >= obs_gm),
    stringsAsFactors = FALSE
  )
}

parallel::stopCluster(cl_comp)

trunc_genes <- function(gs) {
  if (length(gs) <= 10) paste(gs, collapse = ", ")
  else paste0(paste(gs[1:10], collapse = ", "), ", ...")
}

component_loo_tbl <- component_loo_raw %>%
  mutate(
    comp_genes          = vapply(comp_genes_raw, trunc_genes, character(1)),
    p_value_full        = p_value_full,
    delta_p             = p_value_without - p_value_full,
    significant_without = p_value_without < 0.05
  ) %>%
  select(comp_id, comp_size, comp_genes, p_value_without,
         p_value_full, delta_p, significant_without) %>%
  arrange(desc(delta_p))

component_loo_tbl %>%
  knitr::kable(digits = 4,
               caption = "Component LOO: p-value change when each component is removed")
```

```{r}
#| label: plot-component-loo
#| fig-width: 12
#| fig-height: 6
comp_plot_df <- component_loo_tbl %>%
  mutate(comp_label = paste0("C", comp_id, " (n=", comp_size, ")"),
         comp_label = factor(comp_label, levels = comp_label[order(-delta_p)]))

ggplot(comp_plot_df, aes(x = comp_label, y = delta_p, fill = comp_size)) +
  geom_col(width = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_viridis_c(option = "D") +
  labs(
    title = "Component LOO: contribution to Global Moran's I signal",
    subtitle = paste0("delta_p = p(without component) - p(full set); baseline p = ",
                      signif(p_value_full, 3)),
    x    = "Component",
    y    = expression(Delta * "p = p(without) - p(full)"),
    fill = "Component\nsize"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Gene-Level Leave-One-Out (Jackknife)

For each GWAS gene on the network, remove that single gene and rerun Global Moran's I (bins = 10, same permutations). This jackknife gives per-gene contribution to the global signal, including singletons as a sanity check.

```{r}
#| label: gene-loo
#| cache: true

library(foreach)
library(doParallel)

cl_gene <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl_gene)
parallel::clusterEvalQ(cl_gene, {
  suppressPackageStartupMessages({ library(igraph); library(Matrix) })
  NULL
})
parallel::clusterExport(
  cl_gene,
  varlist = c("A_loo", "nodes_loo", "bin_cache", "n_perm_loo",
              "gwas_on_net",
              "global_moran_I", "sample_degree_matched"),
  envir = environment()
)

gene_loo_raw <- foreach(gg = gwas_on_net, .combine = bind_rows) %dopar% {
  reduced <- setdiff(gwas_on_net, gg)

  if (length(reduced) < 3) {
    return(data.frame(gene = gg, p_value_without = NA_real_,
                      stringsAsFactors = FALSE))
  }

  idx_obs <- match(reduced, nodes_loo)
  x_obs   <- integer(length(nodes_loo)); x_obs[idx_obs] <- 1L
  obs_gm  <- global_moran_I(A_loo, x_obs)

  set.seed(42L + which(gwas_on_net == gg)[1])
  rand_gm <- numeric(n_perm_loo)
  for (i in seq_len(n_perm_loo)) {
    s     <- sample_degree_matched(reduced, bin_cache, num_bins = 10, replace = TRUE)
    idx_s <- match(s, nodes_loo)
    x_s   <- integer(length(nodes_loo)); x_s[idx_s] <- 1L
    rand_gm[i] <- global_moran_I(A_loo, x_s)
  }

  data.frame(gene = gg, p_value_without = mean(rand_gm >= obs_gm),
             stringsAsFactors = FALSE)
}

parallel::stopCluster(cl_gene)

# Annotate with subgraph degree, network degree, component, articulation status
gene_loo_tbl <- gene_loo_raw %>%
  left_join(
    tibble(
      gene            = V(gwas_subg)$name,
      subgraph_degree = degree(gwas_subg),
      network_degree  = degree(largest_subgraph_undirected, v = V(gwas_subg)$name),
      comp_id         = comp$membership,
      comp_size       = comp$csize[comp$membership],
      is_articulation_point = V(gwas_subg)$name %in% art_names
    ),
    by = "gene"
  ) %>%
  mutate(
    p_value_full        = p_value_full,
    delta_p             = p_value_without - p_value_full,
    significant_without = p_value_without < 0.05
  ) %>%
  select(gene, subgraph_degree, network_degree, comp_id, comp_size,
         is_articulation_point, p_value_without, p_value_full,
         delta_p, significant_without) %>%
  arrange(desc(delta_p))
```

```{r}
#| label: gene-loo-top20
gene_loo_tbl %>%
  head(20) %>%
  select(gene, subgraph_degree, network_degree, comp_id, comp_size,
         is_articulation_point, p_value_without, delta_p, significant_without) %>%
  knitr::kable(digits = 4,
               caption = "Top 20 genes by LOO contribution (delta_p descending)")
```

```{r}
#| label: gene-loo-datatable
DT::datatable(
  gene_loo_tbl,
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Full gene-level LOO table (browsable)"
) %>%
  DT::formatRound(columns = c("p_value_without", "p_value_full", "delta_p"),
                  digits = 4)
```

```{r}
#| label: plot-gene-loo-scatter
#| fig-width: 10
#| fig-height: 7
scatter_df <- gene_loo_tbl %>%
  mutate(comp_label = ifelse(comp_size == 1, "Singleton",
                             paste0("C", comp_id)))

ggplot(scatter_df, aes(x = subgraph_degree, y = delta_p, color = comp_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 2.2, alpha = 0.8) +
  ggrepel::geom_text_repel(
    data = scatter_df %>% filter(is_articulation_point),
    aes(label = gene),
    size = 3, max.overlaps = 30, show.legend = FALSE
  ) +
  labs(
    title = "Gene LOO: degree vs. contribution to Global Moran's I",
    subtitle = "Articulation points labeled; color = component membership",
    x     = "Subgraph degree",
    y     = expression(Delta * "p = p(without) - p(full)"),
    color = "Component"
  ) +
  theme_minimal(base_size = 12)
```

```{r}
#| label: plot-gene-loo-top20-bar
#| fig-width: 12
#| fig-height: 6
top20_df <- gene_loo_tbl %>%
  head(20) %>%
  mutate(gene = factor(gene, levels = gene[order(-delta_p)]))

ggplot(top20_df, aes(x = gene, y = delta_p, fill = is_articulation_point)) +
  geom_col(width = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
                    labels = c("FALSE" = "No", "TRUE" = "Yes")) +
  labs(
    title = "Top 20 genes by LOO contribution",
    x     = "Gene",
    y     = expression(Delta * "p = p(without) - p(full)"),
    fill  = "Articulation\npoint"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: plot-gene-loo-singleton-hist
#| fig-width: 10
#| fig-height: 5
singleton_df <- gene_loo_tbl %>% filter(comp_size == 1)

sing_mean <- mean(singleton_df$delta_p, na.rm = TRUE)
sing_sd   <- sd(singleton_df$delta_p,   na.rm = TRUE)

ggplot(singleton_df, aes(x = delta_p)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
           label = sprintf("mean = %.4f\nsd = %.4f\nn = %d",
                           sing_mean, sing_sd, nrow(singleton_df))) +
  labs(
    title    = "Singleton LOO delta_p distribution (sanity check)",
    subtitle = "Expect delta_p ~ 0: singletons have no neighbors, so should not contribute to Global Moran's I",
    x        = expression(Delta * "p"),
    y        = "Count"
  ) +
  theme_minimal(base_size = 12)
```

```{r}
#| label: gene-loo-spearman
non_singleton <- gene_loo_tbl %>% filter(comp_size > 1)

sp <- suppressWarnings(cor.test(
  non_singleton$subgraph_degree,
  non_singleton$delta_p,
  method = "spearman"
))

cat("Spearman correlation (subgraph_degree vs delta_p, non-singletons only):\n")
cat("  rho     =", round(sp$estimate, 4), "\n")
cat("  p-value =", format.pval(sp$p.value), "\n")
cat("  n       =", nrow(non_singleton), "\n")
```

# Summary

```{r}
#| label: final-summary
cat("=== CRC GWAS Network Analysis Summary ===\n\n")

cat("Gene set: ", length(gwas_on_net), " genes on network\n")
cat("Degree bias: KS p =", format.pval(ks.test(deg_gwas, deg_all)$p.value), "\n\n")

# Best degree-matched result (bins=4 as default reference)
ref_bins <- 4
ref_row <- stability_df %>% filter(stat == "ICS", ics_bins == ref_bins)
if (nrow(ref_row) > 0) {
  cat("ICS (bins=", ref_bins, "): p =", ref_row$p_value, "\n")
}
ref_row <- stability_df %>% filter(stat == "Global Moran's I", ics_bins == ref_bins)
if (nrow(ref_row) > 0) {
  cat("Global Moran (bins=", ref_bins, "): p =", ref_row$p_value, "\n")
}
ref_row <- stability_df %>% filter(stat == "Join count", ics_bins == ref_bins)
if (nrow(ref_row) > 0) {
  cat("Join count (bins=", ref_bins, "): p =", ref_row$p_value, "\n")
}

cat("\nSubgraph structure:\n")
cat("  Connected components:", comp$no, "\n")
cat("  Largest component:", max(comp$csize), "genes\n")
cat("  Articulation points:", length(art_names), "/", length(gwas_on_net), "\n")
if (nrow(art_impact) > 0) {
  cat("  Max component increase on removal:",
      max(art_impact$delta_components), "(gene:",
      art_impact$gene[which.max(art_impact$delta_components)], ")\n")
}
```
