---
title: "Degree-Typical Signal: Cost of Degree Control"
format: html
editor: visual
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(igraph)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(foreach)
  library(doParallel)
  library(here)
})

source(here::here("R/stats.R"))
source(here::here("R/sampling.R"))
source(here::here("R/auc.R"))
source(here::here("R/simulation_synthetic.R"))
```

```{r}
#| label: load-data
largest_subgraph_undirected <- readRDS(here::here("networks", "largest_subgraph_undirected.rds"))
bin_cache                   <- readRDS(here::here("networks", "degree_bin_cache.rds"))
```

# Design

This notebook tests whether degree-matched null models cost statistical power when the signal genes are **degree-typical** — i.e., their degree distribution matches the network background. If the signal is degree-typical, a degree-matched null should be no harder to beat than a uniform null, and Global Moran's I should show similar AUC under both.

We use `plant_degree_matched_cluster()` (Scenario 7) to generate a BFS cluster whose degree distribution is not significantly different from the full network (KS test p \> 0.3). We then run a standard AUC framework varying signal fraction and set size K.

# Parameters

```{r}
#| label: params
signal_frac_vec <- c(0.05, 0.10, 0.20, 0.40, 0.60)
K_vec           <- c(100, 200, 500)
reps            <- 100
n_perm          <- 1000
cluster_size    <- 50
n_cores         <- 7
seed_base       <- 777
```

# Plant degree-matched cluster

```{r}
#| label: plant-cluster
#| cache: true
set.seed(seed_base)
s7 <- plant_degree_matched_cluster(
  largest_subgraph_undirected,
  cluster_size = cluster_size,
  max_attempts = 200
)

cat("Signal genes:", length(s7$signal_genes), "\n")
cat("KS p-value (cluster vs network degree):", round(s7$ks_pvalue, 4), "\n")
cat("Cluster degree: mean =", round(mean(s7$cluster_degs), 1),
    ", median =", median(s7$cluster_degs), "\n")
cat("Network degree: mean =", round(mean(s7$network_degs), 1),
    ", median =", median(s7$network_degs), "\n")
```

# Feasibility

```{r}
#| label: feasibility
n_signal <- length(s7$signal_genes)

feasibility <- expand.grid(K = K_vec, signal_frac = signal_frac_vec,
                           stringsAsFactors = FALSE) %>%
  mutate(
    signal_needed = ceiling(K * signal_frac),
    feasible      = signal_needed <= n_signal
  )

feasibility %>%
  pivot_wider(names_from = signal_frac, values_from = c(signal_needed, feasible),
              names_glue = "frac={signal_frac}_{.value}") %>%
  knitr::kable(caption = paste0(
    "Feasibility: signal_needed <= ", n_signal, " planted genes"
  ))

feasible_combos <- feasibility %>% filter(feasible)
cat("\nFeasible combinations:", nrow(feasible_combos), "/", nrow(feasibility), "\n")

if (any(!feasibility$feasible)) {
  cat("SKIPPING:\n")
  feasibility %>%
    filter(!feasible) %>%
    mutate(msg = paste0("  K=", K, " frac=", signal_frac,
                        " (need ", signal_needed, " > ", n_signal, " available)")) %>%
    pull(msg) %>%
    cat(sep = "\n")
}
```

# AUC engine

```{r}
#| label: auc-engine
run_degree_typical_auc <- function(
  graph, signal_genes, K, signal_frac_vec, reps, n_perm,
  bin_cache, seed, n_cores
) {
  prep  <- prep_graph_fast(graph)
  g     <- prep$g
  A     <- prep$A
  nodes <- prep$nodes

  signal_genes <- intersect(signal_genes, nodes)
  all_genes    <- nodes

  # Only bins=1 (uniform) and bins=10 (degree-matched) for Global Moran comparison
  ics_bins_set <- c(1, 10)

  task_df <- expand.grid(
    rep         = seq_len(reps),
    signal_frac = signal_frac_vec,
    label       = c(1L, 0L),
    stringsAsFactors = FALSE
  ) %>%
    mutate(task_id = row_number())

  cl <- parallel::makeCluster(n_cores)
  on.exit(parallel::stopCluster(cl), add = TRUE)
  doParallel::registerDoParallel(cl)

  parallel::clusterEvalQ(cl, {
    suppressPackageStartupMessages({ library(igraph); library(Matrix) })
    NULL
  })

  parallel::clusterExport(
    cl,
    varlist = c(
      "g", "A", "nodes", "all_genes", "signal_genes",
      "K", "n_perm", "ics_bins_set", "seed",
      "bin_cache",
      "fast_joincount", "global_moran_I", "ics_norm",
      "sample_degree_matched"
    ),
    envir = environment()
  )

  res_long <- foreach(
    ii = seq_len(nrow(task_df)),
    .combine = dplyr::bind_rows,
    .packages = c("igraph", "Matrix", "dplyr")
  ) %dopar% {

    rep_id <- task_df$rep[ii]
    frac   <- task_df$signal_frac[ii]
    label  <- task_df$label[ii]

    set.seed(seed + ii)

    s <- as.integer(round(K * frac))
    if (s < 1) s <- 1
    if (s > K) s <- K

    if (label == 1L) {
      sig_core <- sample(signal_genes, min(s, length(signal_genes)), replace = FALSE)
      fillers  <- if (K - length(sig_core) > 0) {
        sample(setdiff(all_genes, sig_core), K - length(sig_core), replace = FALSE)
      } else {
        character(0)
      }
      target_nodes <- c(sig_core, fillers)
    } else {
      target_nodes <- sample(all_genes, K, replace = FALSE)
    }

    idx_target <- match(target_nodes, nodes)

    x_obs   <- integer(length(nodes)); x_obs[idx_target] <- 1L
    obs_gm  <- global_moran_I(A, x_obs)
    obs_jc  <- fast_joincount(A, x_obs)
    obs_ics <- ics_norm(induced_subgraph(g, vids = target_nodes))

    all_out <- list()

    for (b in ics_bins_set) {
      rand_ics <- numeric(n_perm)
      rand_gm  <- numeric(n_perm)
      rand_jc  <- numeric(n_perm)

      for (p in seq_len(n_perm)) {
        sset <- if (b > 1) {
          sample_degree_matched(target_nodes, bin_cache, num_bins = b, replace = TRUE)
        } else {
          sample(all_genes, K, replace = TRUE)
        }
        idx_s <- match(sset, nodes)
        x_s   <- integer(length(nodes)); x_s[idx_s] <- 1L

        rand_ics[p] <- ics_norm(induced_subgraph(g, vids = sset))
        rand_gm[p]  <- global_moran_I(A, x_s)
        rand_jc[p]  <- fast_joincount(A, x_s)
      }

      null_label <- if (b > 1) paste0("dm_bins=", b) else "uniform"

      all_out[[length(all_out) + 1]] <- c(
        setNames(mean(rand_ics <= obs_ics), paste0("ICS_bins=", b)),
        setNames(mean(rand_gm >= obs_gm),  paste0("GlobalMoran_", null_label)),
        setNames(mean(rand_jc >= obs_jc),  paste0("Joincount_", null_label))
      )
    }

    out <- unlist(all_out)

    data.frame(
      rep         = rep_id,
      signal_frac = frac,
      label       = label,
      method      = names(out),
      p_value     = as.numeric(out),
      stringsAsFactors = FALSE
    )
  }

  res_long %>%
    mutate(score = -log10(pmax(p_value, 1e-300)))
}
```

# Run AUC across K values

```{r}
#| label: run-auc-v2
#| cache: true
feasible_K <- feasible_combos %>%
  group_by(K) %>%
  summarise(max_frac = max(signal_frac), .groups = "drop") %>%
  pull(K) %>%
  unique()

all_res <- bind_rows(lapply(feasible_K, function(k) {
  # Only use signal fractions feasible for this K
  ok_fracs <- feasible_combos %>% filter(K == k) %>% pull(signal_frac)
  cat("K =", k, ", signal fractions:", paste(ok_fracs, collapse = ", "), "\n")

  res <- run_degree_typical_auc(
    graph           = largest_subgraph_undirected,
    signal_genes    = s7$signal_genes,
    K               = k,
    signal_frac_vec = ok_fracs,
    reps            = reps,
    n_perm          = n_perm,
    bin_cache       = bin_cache,
    seed            = seed_base + k,
    n_cores         = n_cores
  )
  res %>% mutate(K = k)
}))

cat("Total rows:", nrow(all_res), "\n")
```

# Compute AUC

```{r}
#| label: compute-auc
auc_all <- all_res %>%
  group_by(K, signal_frac, method) %>%
  group_modify(~ {
    ra <- roc_auc_from_p(.x, score_col = "score", label_col = "label")
    tibble(auc = ra$auc)
  }) %>%
  ungroup()
```

# Cost of degree control: Global Moran uniform vs bins=10

```{r}
#| label: cost-table
cost_df <- auc_all %>%
  filter(method %in% c("GlobalMoran_uniform", "GlobalMoran_dm_bins=10")) %>%
  select(K, signal_frac, method, auc) %>%
  pivot_wider(names_from = method, values_from = auc) %>%
  rename(
    auc_uniform = GlobalMoran_uniform,
    auc_bins10  = `GlobalMoran_dm_bins=10`
  ) %>%
  mutate(delta = auc_uniform - auc_bins10)

cost_df %>%
  knitr::kable(digits = 3,
               caption = "Global Moran's I AUC: uniform vs bins=10 for degree-typical signal (delta = cost of degree control)")
```

## Cost interpretation

For degree-typical signal, we expect delta to be near zero: degree-matched nulls should not cost power when the signal genes have the same degree distribution as the network background. Large positive delta would indicate that even innocuous degree structure inflates apparent clustering under uniform null.

```{r}
#| label: cost-summary
cat("Mean delta across all conditions:", round(mean(cost_df$delta, na.rm = TRUE), 4), "\n")
cat("Max delta:", round(max(cost_df$delta, na.rm = TRUE), 4), "\n")
cat("Min delta:", round(min(cost_df$delta, na.rm = TRUE), 4), "\n")

if (all(abs(cost_df$delta) < 0.05, na.rm = TRUE)) {
  cat("\nAll deltas < 0.05 — degree control has negligible cost for degree-typical signal.\n")
} else {
  cat("\nSome deltas >= 0.05 — degree control may affect power even for degree-typical signal.\n")
}
```

# Cost of degree control: Join Count uniform vs bins=10

```{r}
#| label: jc-cost-table
jc_cost <- auc_all %>%
  filter(method %in% c("Joincount_uniform", "Joincount_dm_bins=10")) %>%
  select(K, signal_frac, method, auc) %>%
  pivot_wider(names_from = method, values_from = auc) %>%
  rename(auc_uniform = Joincount_uniform, auc_bins10 = `Joincount_dm_bins=10`) %>%
  mutate(delta = auc_uniform - auc_bins10)

jc_cost %>%
  knitr::kable(digits = 3,
               caption = "Join count AUC: uniform vs bins=10 for degree-typical signal (delta = cost of degree control)")
```

```{r}
#| label: jc-cost-summary
cat("Join Count — Mean delta:", round(mean(jc_cost$delta, na.rm = TRUE), 4), "\n")
cat("Max delta:", round(max(jc_cost$delta, na.rm = TRUE), 4), "\n")
cat("Min delta:", round(min(jc_cost$delta, na.rm = TRUE), 4), "\n")
```

# Plots

## AUC vs signal fraction, faceted by K

```{r}
#| label: plot-auc-vs-frac
#| fig-width: 14
#| fig-height: 5
ggplot(auc_all, aes(x = signal_frac, y = auc, color = method, group = method)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ paste0("K = ", K), nrow = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Degree-typical signal: AUC vs signal fraction",
    subtitle = paste0("S7 cluster (", length(s7$signal_genes), " genes, KS p=",
                      round(s7$ks_pvalue, 3), ")"),
    x     = "Signal fraction",
    y     = "AUC",
    color = "Method"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8),
        strip.text = element_text(face = "bold")) +
  guides(color = guide_legend(ncol = 3))
```

## Global Moran: uniform vs bins=10 side-by-side

```{r}
#| label: plot-gmoran-comparison
#| fig-width: 12
#| fig-height: 5
gmoran_df <- auc_all %>%
  filter(grepl("^GlobalMoran", method))

ggplot(gmoran_df, aes(x = signal_frac, y = auc, color = method, group = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ paste0("K = ", K), nrow = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_manual(values = c("GlobalMoran_uniform" = "#E69F00",
                                "GlobalMoran_dm_bins=10" = "#0072B2")) +
  labs(
    title = "Global Moran's I: cost of degree control for degree-typical signal",
    subtitle = "Lines should overlap if degree control has no cost",
    x     = "Signal fraction",
    y     = "AUC",
    color = "Null model"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

## Delta (cost) heatmap

```{r}
#| label: plot-cost-heatmap
#| fig-width: 8
#| fig-height: 4
ggplot(cost_df, aes(x = factor(signal_frac), y = factor(K), fill = delta)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(delta, 3)), size = 3.5) +
  scale_fill_gradient2(low = "#0072B2", mid = "white", high = "#D55E00",
                       midpoint = 0, limits = c(-0.15, 0.15)) +
  labs(
    title = "Cost of degree control (AUC_uniform - AUC_bins=10)",
    subtitle = "Near zero = degree control is free for degree-typical signal",
    x    = "Signal fraction",
    y    = "K (set size)",
    fill = "Delta AUC"
  ) +
  theme_minimal(base_size = 12)
```

## Global Moran vs Join Count cost comparison

```{r}
#| label: plot-cost-comparison
#| fig-width: 14
#| fig-height: 5
cost_combined <- bind_rows(
  cost_df %>% mutate(statistic = "Global Moran"),
  jc_cost %>% mutate(statistic = "Join Count")
)

ggplot(cost_combined, aes(x = signal_frac, y = delta,
                           color = statistic, group = statistic)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap(~ paste0("K = ", K), nrow = 1) +
  scale_color_manual(values = c("Global Moran" = "#E69F00", "Join Count" = "#A65628")) +
  labs(
    title = "Cost of degree control: Global Moran vs Join Count",
    subtitle = "Delta = AUC_uniform - AUC_bins=10 (positive = uniform has higher AUC)",
    x     = "Signal fraction",
    y     = "Delta AUC",
    color = "Statistic"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

# Save results

```{r}
#| label: save
dir.create(here::here("results"), showWarnings = FALSE)
saveRDS(all_res,  here::here("results", "degree_typical_auc_raw.rds"))
saveRDS(auc_all,  here::here("results", "degree_typical_auc_table.rds"))
saveRDS(cost_df,  here::here("results", "degree_typical_cost_table.rds"))
saveRDS(jc_cost,  here::here("results", "degree_typical_jc_cost_table.rds"))
```
